---
title: "Retail Sales Data with Seasonal Trends & Marketing"
subtitle: "Project Proposal — Group 1"
author: "Jake Evans · Mickyas Shawel · Aldo Naranjo · Luis Berumen · Tim Arana"
institute: "IBM 6540 — Applied Machine Learning for Marketing | Cal Poly Pomona"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    embed-resources: true
    footer: "IBM 6540 — Group 1 Project Proposal | Summer 2026"
    toc: false
---

## Group 1

| Member         | RO Lead |
|----------------|---------|
| Jake Evans     | RO \_\_ |
| Mickyas Shawel | RO \_\_ |
| Aldo Naranjo   | RO \_\_ |
| Luis Berumen   | RO \_\_ |
| Tim Arana      | RO \_\_ |

::: aside
Each member leads one Research Objective (fill in assignments above).
:::

## Data & Repository

- **Dataset:** Retail Sales Data with Seasonal Trends & Marketing (Kaggle)
  - <https://www.kaggle.com/datasets/abdullah0a/retail-sales-data-with-seasonal-trends-and-marketing>
  - 30,000 records of transaction-level retail sales
- **Group Repo:** <https://github.com/mjshawell/IBM-6540-Group-Project>

# Ch1: Introduction {background-color="#23351F"}

## Problem Definition

::: callout-note
## Group: draft together

- Who is the audience/client? (e.g., a multi-location retail chain's marketing team)
- What decision are they struggling with? (e.g., when to spend on marketing, how deep to discount, how to plan for seasonal demand)
- One or two sentences framing the business problem.
:::

## Research Objectives

- **RO1 — Seasonality:** How do sales revenue and units sold vary across seasons and months?
- **RO2 — Marketing:** What is the relationship between marketing spend and sales revenue?
- **RO3 — Discounting:** How does discount percentage affect units sold and revenue?
- **RO4 — Holidays & Timing:** How do holiday periods and day of the week influence sales?
- **RO5 — Category & Location:** How does sales performance differ by product category and store location?

::: aside
*Draft ROs — group to revise wording and confirm one lead per RO.*
:::

## Importance of the Topic

::: callout-note
## Group: draft together

Why does this matter for the audience? Starting angles:

- Seasonal planning affects inventory, staffing, and budgets
- Marketing dollars need ROI justification
- Discounts drive volume but can erode margin
- Findings translate into an actionable promotion calendar
:::

# Ch3: Methods {background-color="#23351F"}

## Population

::: callout-note
## Group: define

The population the sample represents — e.g., retail sales transactions across multiple stores, locations, and product categories. Note that this is a synthetic/educational dataset and state what real-world population it stands in for.
:::

## Sampling / Data Collection

- Secondary data: public Kaggle dataset (link on slide 2)
- 30,000 transaction records
- **Key variables:**
  - Sales Revenue (USD), Units Sold
  - Discount Percentage, Marketing Spend (USD)
  - Date, Day of the Week, Holiday Effect
  - Store ID, Store Location, Product Category

## Data Wrangling

- **From records to data:** import CSV with `readr::read_csv()`; store in repo `data/` folder
- **Tidying:** clean column names (`janitor::clean_names()`); check missing values and duplicates; convert `Date` to date type; make `Holiday Effect`, `Day of the Week`, `Product Category`, `Store Location` factors
- **Transforming:** derive `month`, `season`, and `year` from `Date`; create spend/discount tiers as needed per RO

# Research Objectives {background-color="#23351F"}

## RO1: Discounts & Units Sold — Lead: Jake Evans

- **Objective:** Analyze whether higher discount percentages are associated with higher units sold
- **IV:** Discount Percentage · **DV:** Units Sold
- **Controls:** Product Category, Store Location, Day of Week, Holiday Effect
- **Wrangling:** clean discount variable; group into Low / Moderate / High; summarize units sold by group
- **Hypothesis direction:** higher discount percentages will be associated with higher units sold

```{r}
#| label: ro1-setup
#| include: false
library(tidyverse)
library(janitor)
library(lubridate)
library(scales)
library(knitr)
theme_set(theme_minimal(base_size = 14))

retail <- read_csv("retail_sales_data.csv") |>
  clean_names()

retail_clean <- retail |>
  mutate(
    date = as.Date(date),
    discount_group = case_when(
      discount_percentage <= 10 ~ "Low discount",
      discount_percentage <= 25 ~ "Moderate discount",
      TRUE                      ~ "High discount"
    ),
    discount_group = factor(
      discount_group,
      levels = c("Low discount", "Moderate discount", "High discount")
    )
  )

ro1_summary <- retail_clean |>
  group_by(discount_group) |>
  summarize(
    n = n(),
    avg_units_sold = mean(units_sold, na.rm = TRUE),
    avg_revenue    = mean(sales_revenue_usd, na.rm = TRUE),
    .groups = "drop"
  )
```

## RO1: Discount % vs. Units Sold

```{r}
#| label: ro1-scatter
#| code-fold: true
#| fig-height: 4.5
ggplot(retail_clean, aes(discount_percentage, units_sold)) +
  geom_point(alpha = 0.08, color = "#23351F") +
  geom_smooth(method = "lm", color = "#D9A441", linewidth = 1.2) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(x = "Discount Percentage", y = "Units Sold")
```

## RO1: Units Sold by Discount Group

```{r}
#| label: ro1-boxplot
#| code-fold: true
#| fig-height: 4.5
ggplot(retail_clean, aes(discount_group, units_sold, fill = discount_group)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.15) +
  scale_fill_manual(values = c("#4E6E8E", "#7FA05C", "#D9A441")) +
  labs(x = NULL, y = "Units Sold")
```

## RO1: Average Units Sold by Discount Group

```{r}
#| label: ro1-bars
#| code-fold: true
#| fig-height: 4
ggplot(ro1_summary, aes(discount_group, avg_units_sold, fill = discount_group)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(avg_units_sold, 1)), vjust = -0.4) +
  scale_fill_manual(values = c("#4E6E8E", "#7FA05C", "#D9A441")) +
  labs(x = NULL, y = "Average Units Sold")
```

## RO1: Summary by Discount Group

```{r}
#| label: ro1-table
#| code-fold: true
ro1_summary |>
  mutate(
    avg_units_sold = round(avg_units_sold, 2),
    avg_revenue    = dollar(avg_revenue, accuracy = 0.01)
  ) |>
  kable(col.names = c("Discount Group", "Transactions", "Avg Units Sold", "Avg Revenue"))
```

## RO2: Marketing Spend — Lead: \_\_\_\_\_\_

- **IV:** Marketing Spend (USD)
- **DV:** Sales Revenue
- **Wrangling plan:** flag zero-spend vs. positive-spend transactions; create spend tiers
- **Visualization plan:** scatterplot of spend vs. revenue with trend line; tier comparison bars

## RO3: Discounting — Lead: \_\_\_\_\_\_

- **IV:** Discount Percentage
- **DVs:** Units Sold, Sales Revenue
- **Wrangling plan:** bin discounts (none / low / medium / high); compare distributions across bins
- **Visualization plan:** boxplots of units sold by discount bin; discount vs. revenue scatter

## RO4: Holidays & Timing — Lead: \_\_\_\_\_\_

- **IVs:** Holiday Effect, Day of the Week
- **DVs:** Sales Revenue, Units Sold
- **Wrangling plan:** confirm factor levels; order days Mon–Sun; group and summarize by holiday flag and day
- **Visualization plan:** grouped bar chart of average revenue by day, split by holiday vs. non-holiday

## RO5: Category & Location — Lead: \_\_\_\_\_\_

- **IVs:** Product Category, Store Location
- **DVs:** Sales Revenue, Units Sold
- **Wrangling plan:** summarize revenue/units by category and location; identify top and bottom performers
- **Visualization plan:** faceted bar charts by location; category heatmap

# Conclusion {background-color="#23351F"}

## Conclusion

::: callout-note
## Group: draft together

- Restate the problem and the five ROs in one or two lines
- What the client gains: a data-driven view of seasonality, promotion effectiveness, and category/location performance
- Next steps toward Ch4 (Results) and Ch5 (Recommendations)
:::

## Thank You

**Questions?**

Group 1 — IBM 6540 \| Summer 2026

Repo: <https://github.com/mjshawell/IBM-6540-Group-Project>
