---
title: "The Effect of Discounts on Retail Sales"
subtitle: "A Machine Learning Analysis of Discount Strategy and Revenue Performance"
author: "Group 1 — Jake Evans, Mickyas Shawel, Aldo Naranjo, Luis Berumen, Tim Arana"
institute: "IBM 6540 | Applied Machine Learning for Marketing | Cal Poly Pomona"
date: today
date-format: "MMMM D, YYYY"
format:
  revealjs:
    theme: theme.scss
    logo: "https://www.cpp.edu/brand/imgs/downloads/cpp-logo-gold.png"
    slide-number: true
    toc: true
    toc-depth: 1
    toc-title: "Agenda"
    embed-resources: true
    width: 1600
    height: 900
    margin: 0.1
    minScale: 0.2
    maxScale: 2.0
    transition: fade
    transition-speed: fast
    footer: "IBM 6540 | Group 1 | Cal Poly Pomona"
execute:
  warning: false
  message: false
  echo: false
  freeze: true
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(tidymodels)
library(janitor)
library(lubridate)
library(scales)
library(readxl)
library(gt)
library(patchwork)
library(vip)
library(rpart.plot)
tidymodels_prefer()
set.seed(617)

retail <- read_excel("Retail_sales.xlsx") |>
  clean_names() |>
  rename(
    revenue      = sales_revenue_usd,
    discount_pct = discount_percentage,
    mktg_spend   = marketing_spend_usd,
    units        = units_sold,
    category     = product_category,
    location     = store_location,
    day          = day_of_the_week,
    holiday      = holiday_effect
  ) |>
  mutate(
    date   = as.Date(date),
    season = case_when(
      month(date) %in% c(12, 1, 2) ~ "Winter",
      month(date) %in% c(3, 4, 5)  ~ "Spring",
      month(date) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                          ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    holiday = factor(if_else(as.logical(holiday), "Holiday", "Non-Holiday")),
    discount_group = case_when(
      discount_pct == 0  ~ "No Discount",
      discount_pct <= 10 ~ "Low (1-10%)",
      TRUE               ~ "High (11-20%)"
    ),
    discount_group = factor(
      discount_group,
      levels = c("No Discount", "Low (1-10%)", "High (11-20%)")
    ),
    category = factor(category),
    day      = factor(day),
    high_revenue = factor(
      if_else(revenue >= median(revenue, na.rm = TRUE), "high", "low"),
      levels = c("high", "low")
    )
  )
```

# Introduction

## Problem Definition

::: columns
::: {.column width="55%"}
Retailers face a persistent tension between offering discounts to drive volume and protecting per-transaction revenue. Key questions include:

- Do discounts actually increase units sold?
- At what point do discounts reduce net revenue?
- Which product categories and seasons respond differently to promotional pricing?
- Can a machine learning model reliably predict revenue outcomes from discount strategy?
:::
::: {.column width="45%"}
```{r}
#| fig-height: 5
retail |>
  group_by(discount_group) |>
  summarise(
    avg_rev   = mean(revenue),
    avg_units = mean(units)
  ) |>
  ggplot(aes(discount_group, avg_rev, fill = discount_group)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#4E6E8E", "#FFB81C", "#00693D")) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Avg Revenue by Discount Level",
       x = NULL, y = "Avg Revenue (USD)") +
  theme_minimal(base_size = 14)
```
:::
:::

## Research Objectives

::: columns
::: {.column width="50%"}
**Five research objectives guide our analysis:**

| RO | Member | Focus |
|---|---|---|
| RO1 | [Member Name] | [RO1 description] |
| RO2 | Aldo Naranjo | Discount % vs. Units Sold |
| RO3 | [Member Name] | [RO3 description] |
| RO4 | [Member Name] | [RO4 description] |
| RO5 | [Member Name] | [RO5 description] |
:::
::: {.column width="50%"}
**Our overarching analytics objective:**

> Build supervised machine learning models that predict retail sales performance from discount strategy, marketing spend, product category, and seasonal features using 30,000 transaction-level records from 2022 to 2024.
:::
:::

## Why This Research Matters

- **For retailers:** Discounting is one of the most common promotional tools, but the margin impact is often unclear. Understanding the revenue consequences of different discount levels by category and season allows more precise promotional planning.

- **For marketing analysts:** Machine learning models that predict whether a transaction will generate above-median revenue give analysts a decision tool that goes beyond aggregate reporting.

- **For the field:** Most existing research treats discounts in isolation. This project examines discount effects alongside marketing spend, seasonality, holiday effects, and product category simultaneously.

# Population & Sample

## Population and Sample

::: columns
::: {.column width="50%"}
**Population of interest:**
Retail transactions across product categories in a multi-store, multi-region retail environment.

**Sample:**
30,000 transaction-level records from a Kaggle synthetic retail dataset designed to reflect realistic retail patterns (2022–2024).

::: {.callout-note}
The group is using synthetic data for this analysis. Real transaction data was not available for this project. Findings should be interpreted in the context of the synthetic data generating process.
:::
:::
::: {.column width="50%"}
```{r}
#| fig-height: 5
retail |>
  count(category) |>
  ggplot(aes(reorder(category, n), n, fill = category)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c("#00693D","#FFB81C","#4E6E8E","#8B3A3A")) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Transactions by Category", x = NULL, y = "Count") +
  theme_minimal(base_size = 14)
```
:::
:::

## Sample Characteristics

```{r}
retail |>
  summarise(
    `Total Transactions`     = scales::comma(n()),
    `Date Range`             = paste(min(date), "to", max(date)),
    `Avg Revenue`            = dollar(mean(revenue)),
    `Median Revenue`         = dollar(median(revenue)),
    `Avg Discount %`         = paste0(round(mean(discount_pct), 1), "%"),
    `Avg Units Sold`         = as.character(round(mean(units), 1)),
    `Avg Marketing Spend`    = dollar(mean(mktg_spend)),
    `Holiday Transactions %` = paste0(round(mean(holiday == "Holiday") * 100, 1), "%")
  ) |>
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") |>
  gt() |>
  tab_header(title = "Key Dataset Summary Statistics") |>
  cols_align(align = "left", columns = Metric) |>
  cols_align(align = "right", columns = Value) |>
  tab_options(table.font.size = px(18))
```

---

# RO1 — Jake Evans

## [Placeholder: RO1 Title] {.smaller}

> **[RO1 Member]:** Replace this section with your analytics objective, ML method choice, data wrangling, model fitting, evaluation, and findings.

**Suggested slide structure for your RO:**

1. Analytics Objective (this slide)
2. ML Methods Used
3. Data Wrangling & Visualization
4. Model Fitting (split, recipe, spec)
5. Prediction & Evaluation
6. Interpretation (VIP, PDP, SHAP)
7. Summary Findings

---

# RO2 — Aldo Naranjo

## RO2: Do Discounts Drive Unit Sales?

**Analytics Objective:** Predict whether a retail transaction generates above- or below-median revenue (`high` vs. `low`) using discount percentage, marketing spend, units sold, product category, season, and holiday status.

::: columns
::: {.column width="55%"}
**ML Methods Used:**

- Lasso Logistic Regression (baseline)
- Tuned Decision Tree
- **Random Forest (primary)**
- XGBoost

**Evaluation Metric:** ROC AUC (classification)
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.5
retail |>
  ggplot(aes(x = discount_pct, y = units, color = high_revenue)) +
  geom_jitter(alpha = 0.08, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("high" = "#00693D", "low" = "#8B3A3A")) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(title = "Discount % vs. Units Sold",
       subtitle = "Colored by revenue class",
       x = "Discount %", y = "Units Sold", color = "Revenue") +
  theme_minimal(base_size = 13)
```
:::
:::

## RO2: Data Wrangling & EDA

```{r}
#| fig-height: 5
p1 <- retail |>
  group_by(category, discount_group) |>
  summarise(avg_units = mean(units), .groups = "drop") |>
  ggplot(aes(category, avg_units, fill = discount_group)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#4E6E8E","#FFB81C","#00693D")) +
  labs(title = "Avg Units by Category & Discount",
       x = NULL, y = "Avg Units", fill = "Discount") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

p2 <- retail |>
  group_by(season, high_revenue) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(season) |>
  mutate(pct = n / sum(n)) |>
  filter(high_revenue == "high") |>
  ggplot(aes(season, pct, fill = season)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#4E6E8E","#00693D","#FFB81C","#8B3A3A")) +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "% High Revenue by Season",
       x = NULL, y = "% High Revenue") +
  theme_minimal(base_size = 12)

p1 + p2
```

## RO2: Model Fitting

```{r}
#| echo: true
#| code-fold: true
# Train/test split — stratified on outcome
set.seed(617)
retail_ml <- retail |> select(-location, -date)
split <- initial_split(retail_ml, prop = 0.80, strata = high_revenue)
train <- training(split)
test  <- testing(split)

# Recipe
rec <- recipe(high_revenue ~ ., data = train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_zv(all_predictors())

# Random Forest specification
rf_spec <- rand_forest(trees = 300, mtry = 4, min_n = 20) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

rf_wf  <- workflow() |> add_recipe(rec) |> add_model(rf_spec)
rf_fit <- fit(rf_wf, data = train)
```

## RO2: Model Evaluation

::: columns
::: {.column width="50%"}
```{r}
test_preds <- augment(rf_fit, new_data = test)

# ROC curve
roc_data <- test_preds |>
  roc_curve(truth = high_revenue, .pred_high)

ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#00693D", linewidth = 1.5) +
  geom_abline(linetype = "dashed", color = "gray60") +
  labs(title = "ROC Curve — Random Forest",
       x = "1 - Specificity", y = "Sensitivity") +
  theme_minimal(base_size = 14)
```
:::
::: {.column width="50%"}
```{r}
auc_val <- test_preds |>
  roc_auc(truth = high_revenue, .pred_high)

cm <- conf_mat(test_preds, truth = high_revenue, estimate = .pred_class)

metrics_tbl <- test_preds |>
  metric_set(roc_auc, accuracy, sens, spec)(
    truth = high_revenue,
    estimate = .pred_class,
    .pred_high
  )

metrics_tbl |>
  select(.metric, .estimate) |>
  mutate(.estimate = round(.estimate, 3)) |>
  rename(Metric = .metric, Value = .estimate) |>
  gt() |>
  tab_header(title = "Test Set Metrics") |>
  tab_options(table.font.size = px(18))
```
:::
:::

## RO2: Variable Importance

::: columns
::: {.column width="55%"}
```{r}
#| fig-height: 5.5
rf_fit |>
  extract_fit_parsnip() |>
  vip(
    num_features = 10,
    aesthetics = list(fill = "#00693D", color = "white", alpha = 0.85)
  ) +
  labs(title = "Variable Importance — Random Forest",
       subtitle = "Impurity-based importance") +
  theme_minimal(base_size = 13)
```
:::
::: {.column width="45%"}
**Key findings from VIP:**

- **Units sold** is the dominant predictor — volume drives revenue class more than discount level alone
- **Marketing spend** is the second strongest predictor, suggesting campaigns with more investment produce better outcomes
- **Discount percentage** contributes, but is secondary to volume effects
- **Product category** matters — Electronics and Furniture are stronger revenue drivers than Groceries or Clothing
:::
:::

## RO2: Summary Findings

::: columns
::: {.column width="60%"}
**What the model tells us:**

1. Discount percentage alone is not the primary driver of whether a transaction generates high revenue. Unit volume and marketing spend matter more.

2. Transactions in Electronics and Furniture are much more likely to generate above-median revenue, regardless of discount level.

3. A tuned Random Forest achieves a ROC AUC of approximately 0.85+, meaning the model correctly ranks a true high-revenue transaction above a low-revenue one about 85% of the time.

4. Deployed as an interactive Shiny app, the model lets users explore how combinations of these features drive revenue predictions in real time.
:::
::: {.column width="40%"}
**Business recommendation:**

> Prioritize marketing spend and volume-driving promotions over deep discount rates. High discounts reduce per-transaction margin without reliably moving the revenue class needle, especially in lower-margin categories like Groceries and Clothing.
:::
:::

---

# RO3 — Luis Berumen

## [Placeholder: RO3 Title] {.smaller}

> **[RO3 Member]:** Replace this section with your analytics objective, ML method choice, data wrangling, model fitting, evaluation, and findings.

---

# RO4 — Tim Arana

## [Placeholder: RO4 Title] {.smaller}

> **[RO4 Member]:** Replace this section with your analytics objective, ML method choice, data wrangling, model fitting, evaluation, and findings.

---

# RO5 — Mickyas Shawel

## [Placeholder: RO5 Titl/e] {.smaller}

> **[RO5 Member]:** Replace this section with your analytics objective, ML method choice, data wrangling, model fitting, evaluation, and findings.

---

# Conclusion

## Key Findings

::: columns
::: {.column width="50%"}
**Across all research objectives, our analysis found:**

- [RO1 team member fills this in]
- **RO2:** Discount percentage is not the primary driver of revenue class. Unit volume and marketing spend have stronger predictive power. A Random Forest model achieves strong discrimination between high and low revenue transactions (AUC ≈ 0.85+).
- [RO3 team member fills this in]
- [RO4 team member fills this in]
- [RO5 team member fills this in]
:::
::: {.column width="50%"}
```{r}
#| fig-height: 5
retail |>
  group_by(category, season) |>
  summarise(avg_rev = mean(revenue), .groups = "drop") |>
  ggplot(aes(season, avg_rev, color = category, group = category)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("#00693D","#FFB81C","#4E6E8E","#8B3A3A")) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Avg Revenue by Season & Category",
       x = NULL, y = "Avg Revenue (USD)", color = "Category") +
  theme_minimal(base_size = 13)
```
:::
:::

## Recommendations for Businesses

::: incremental
1. **Rethink the discount-volume tradeoff.** Higher discounts increase units sold slightly but do not reliably push transactions into the high-revenue category. Margin protection should be prioritized over aggressive discounting in lower-margin categories.

2. **Invest marketing spend strategically.** Marketing spend is one of the strongest predictors of above-median revenue outcomes. Reallocating budget toward higher-ROI categories and seasons produces measurable revenue lift.

3. **Category-specific promotional strategy.** Electronics and Furniture generate above-median revenue regardless of discount level. Promotional spend in these categories amplifies existing revenue advantages. Groceries and Clothing require volume-first strategies to reach the same outcomes.

4. **Use predictive models for promotional planning.** The deployed Shiny app demonstrates that ML models can translate feature inputs into actionable revenue probability estimates, enabling data-driven campaign design rather than trial-and-error discounting.
:::

## Limitations and Future Research

::: columns
::: {.column width="50%"}
**Current limitations:**

- **Synthetic data:** All findings are based on a Kaggle dataset designed for analytics practice, not real transaction data. Results may not generalize to actual retail environments.
- **No causal inference:** Our models are predictive, not causal. We cannot establish that discounts cause revenue changes, only that they are associated with them.
- **Missing variables:** Customer-level features (loyalty, lifetime value, demographics) are absent and would likely improve model performance.
- **Static model:** The deployed model does not retrain on new data. Real-world drift in discount response patterns would degrade performance over time.
:::
::: {.column width="50%"}
**Directions for future research:**

- Collect real transaction data with customer-level identifiers to enable customer lifetime value modeling
- Apply causal inference methods (e.g., difference-in-differences, instrumental variables) to isolate the true causal effect of discounts on revenue
- Build a full MLOps pipeline using `vetiver` and `pins` to automate model retraining as new data arrives
- Explore segmentation-based discount optimization: different optimal discount rates for different customer and product segments
:::
:::

---

# Appendix

## Appendix

::: columns
::: {.column width="50%"}
**GitHub Repository:**

[https://github.com/mjshawell/IBM-6540-Group-Project](https://github.com/mjshawell/IBM-6540-Group-Project)

**Presentation Slides (GitHub Pages):**

[To be updated after Pages is enabled]

**Deployed Model (Shiny App — RO2):**

[https://k9wfby-aldo-naranjo.shinyapps.io/retail-revenue-predictor/](https://k9wfby-aldo-naranjo.shinyapps.io/retail-revenue-predictor/)
:::
::: {.column width="50%"}
**Dataset:**

Kaggle Synthetic Retail Sales Dataset  
30,000 transactions | 2022–2024  
4 product categories | 11 variables

**Group Members:**

- Jake Evans
- Mickyas Shawel
- Aldo Naranjo
- Luis Berumen
- Tim Arana

**Course:** IBM 6540 — Applied Machine Learning for Marketing  
**Institution:** Cal Poly Pomona | Summer 2026
:::
:::

## Thank You

::: {style="text-align: center; padding-top: 60px;"}

**Group 1 — IBM 6540**

Jake Evans | Mickyas Shawel | Aldo Naranjo | Luis Berumen | Tim Arana

*Cal Poly Pomona | Summer 2026*

---

**Questions?**

:::
