---
title: "Analyzing the Effect of Discounts on Retail Sales Volume and Revenue Using Seasonal Retail Data"
subtitle: "Project Proposal — Group 1"
author: "Jake Evans · Mickyas Shawel · Aldo Naranjo · Luis Berumen · Tim Arana"
institute: "IBM 6540 — Applied Machine Learning for Marketing | Cal Poly Pomona"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    width: 1600
    height: 900
    margin: 0.06
    embed-resources: true
    footer: "IBM 6540 — Group 1 Project Proposal | Summer 2026"
    toc: false
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
/* Presentation-wide size and spacing fixes */
.reveal {
  font-size: 29px;
}

.reveal .slides {
  text-align: left;
}

.reveal .slides section {
  padding: 18px 34px;
  box-sizing: border-box;
}

.reveal h1 {
  font-size: 1.65em;
  line-height: 1.08;
  margin-bottom: 0.35em;
}

.reveal h2 {
  font-size: 1.25em;
  line-height: 1.12;
  margin-bottom: 0.45em;
}

.reveal h3 {
  font-size: 1.05em;
  line-height: 1.12;
}

.reveal p,
.reveal li {
  font-size: 0.78em;
  line-height: 1.25;
}

.reveal ul,
.reveal ol {
  margin-top: 0.25em;
  margin-bottom: 0.25em;
}

.reveal li {
  margin-bottom: 0.22em;
}

.reveal table {
  font-size: 0.55em;
  line-height: 1.15;
  width: 96%;
  margin-left: auto;
  margin-right: auto;
}

.reveal th,
.reveal td {
  padding: 0.22em 0.35em;
}

.reveal pre code {
  font-size: 0.62em;
  line-height: 1.15;
}

.reveal .cell-output-display,
.reveal .cell-output {
  margin-top: 0.25em;
}

.reveal .footer {
  font-size: 0.38em;
  bottom: 8px;
}

.reveal .slide-number {
  font-size: 0.45em;
}

.reveal aside,
.reveal .aside {
  font-size: 0.58em;
  line-height: 1.2;
}

/* Title slide specific fixes */
#title-slide {
  text-align: center;
  padding-top: 12px;
}

#title-slide .title {
  font-size: 1.55em !important;
  line-height: 1.02;
  max-width: 1080px;
  margin: 0 auto 0.25em auto;
}

#title-slide .subtitle {
  font-size: 0.75em !important;
  line-height: 1.1;
  margin-bottom: 0.45em;
}

#title-slide .author {
  font-size: 0.62em !important;
  line-height: 1.12;
  max-width: 980px;
  margin: 0 auto 0.3em auto;
}

#title-slide .institute,
#title-slide .date {
  font-size: 0.47em !important;
  line-height: 1.08;
  margin-top: 0.15em;
}

/* Better fit for dense plan slides */
.reveal section:has(ul) h2 + ul li {
  font-size: 0.74em;
}

/* Keep chart slides compact without shrinking axis text too far */
.reveal img,
.reveal svg {
  max-height: 520px;
}
</style>
```

```{r}
#| label: global-setup
#| include: false
library(tidyverse)
library(janitor)
library(lubridate)
library(scales)
library(knitr)
theme_set(theme_minimal(base_size = 12))
knitr::opts_chunk$set(
  fig.width = 8.2,
  fig.height = 4.0,
  out.width = "92%"
)

csv_files <- list.files(".", pattern = "\\.csv$", recursive = TRUE, full.names = TRUE)
data_path <- csv_files[grepl("retail|sales", csv_files, ignore.case = TRUE)][1]
if (is.na(data_path)) data_path <- csv_files[1]

retail <- read_csv(data_path) |>
  clean_names() |>
  mutate(
    date       = as.Date(date),
    month      = month(date, label = TRUE, abbr = TRUE),
    year       = year(date),
    year_month = floor_date(date, "month"),
    season = case_when(
      month(date) %in% c(12, 1, 2)  ~ "Winter",
      month(date) %in% c(3, 4, 5)   ~ "Spring",
      month(date) %in% c(6, 7, 8)   ~ "Summer",
      TRUE                          ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    holiday = if_else(as.logical(holiday_effect), "Holiday", "Non-holiday"),
    day = factor(day_of_the_week,
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday",
                 "Friday", "Saturday", "Sunday")),
    discount_group = case_when(
      discount_percentage == 0  ~ "No discount",
      discount_percentage <= 10 ~ "Low (1–10%)",
      TRUE                      ~ "High (11–20%)"
    ),
    discount_group = factor(discount_group,
      levels = c("No discount", "Low (1–10%)", "High (11–20%)")),
    marketing_spend_group = ntile(marketing_spend_usd, 3),
    marketing_spend_group = case_when(
      marketing_spend_group == 1 ~ "Low marketing spend",
      marketing_spend_group == 2 ~ "Medium marketing spend",
      marketing_spend_group == 3 ~ "High marketing spend"
    ),
    marketing_spend_group = factor(
      marketing_spend_group,
      levels = c("Low marketing spend", "Medium marketing spend", "High marketing spend")
    )
  )
```

## Project Overview

**Main research question:** How does discount percentage affect retail
sales volume and revenue when accounting for product category, marketing
spend, holidays, and seasonal trends?

| Member         | RO Lead |
|----------------|---------|
| Jake Evans     | RO 1    |
| Mickyas Shawel | RO 5    |
| Aldo Naranjo   | RO 2    |
| Luis Berumen   | RO 3    |
| Tim Arana      | RO 4    |

::: aside
Dataset: [Retail Sales Data with Seasonal Trends & Marketing
(Kaggle)](https://www.kaggle.com/datasets/abdullah0a/retail-sales-data-with-seasonal-trends-and-marketing)
· Repo: <https://github.com/mjshawell/IBM-6540-Group-Project>
:::

# Ch1: Introduction {background-color="#23351F"}

## Problem Definition

- **Audience:** retail managers, ecommerce teams, marketing analysts,
  and promotion planners who need to evaluate discount strategy
- Retailers frequently use discounts to attract customers, increase
  demand, and clear inventory
- **The tradeoff:** discounts may increase units sold while reducing
  revenue per unit
- Without data analysis, retailers may overuse discounts, underuse them,
  or apply the wrong discount levels to the wrong products or time
  periods

## Research Objectives

- **RO1 — Seasonality & Timing:** How do sales revenue and units sold
  vary across seasons, months, and holiday periods?
- **RO2 — Discounts & Volume:** Are higher discount percentages
  associated with higher units sold?
- **RO3 — Discounts & Revenue:** Do discounts increase sales revenue?
- **RO4 — Category Differences:** Does the effect of discounts differ
  across product categories?
- **RO5 — Marketing Interaction:** Does marketing spend strengthen the
  impact of discounts?

::: aside
Each member leads one RO — confirm assignments on slide 2.
:::

## Importance of the Topic

- Helps retailers decide when discounts are useful versus when they
  reduce revenue efficiency
- Supports product-level promotional planning by identifying categories
  that respond well to discounts
- Improves marketing strategy by showing whether discounts perform
  better with higher marketing spend
- Supports seasonal planning by comparing holiday and non-holiday
  performance
- Creates a data-driven foundation for pricing, promotion, and revenue
  optimization decisions

## Background & Hypothesis Directions

| RO | Hypothesis Direction |
|-------------------|-----------------------------------------------------|
| RO1 | Revenue and units will vary by season, with holiday periods outperforming regular periods |
| RO2 | Higher discount percentages will be associated with higher units sold |
| RO3 | Discounts may lift revenue up to a point, but higher discounts may not produce proportional gains |
| RO4 | The discount–sales relationship will vary by product category |
| RO5 | Discounts paired with higher marketing spend will show stronger sales outcomes |

# Ch3: Methods {background-color="#23351F"}

## Dataset Overview & Key Variables

- **Source:** Kaggle secondary dataset (synthetic, designed for
  analytics practice) — 30,000 transaction-level records, 2022–2024
- **DVs:** Sales Revenue (USD), Units Sold
- **Primary IV:** Discount Percentage (0–20%)
- **Additional predictors:** Marketing Spend (USD), Product Category,
  Store Location, Day of the Week, Holiday Effect, Date

## Population, Sampling & Data Collection

- **Population represented:** transaction-level retail sales records
  across stores, locations, and product categories
- The dataset is synthetic; it stands in for multi-location retail
  chains that run discount and marketing programs
- **Sampling:** secondary data collection — records are already
  collected and will be prepared for analysis in R

## Data Wrangling Plan

- **From records to data:** import CSV with `readr::read_csv()`; each
  row = one retail sales record
- **Tidying:** `janitor::clean_names()`; check missing values,
  duplicates, and outliers; convert `Date` to date type; standardize
  categorical variables (category, location, day, holiday)
- **Transforming:**
  - `month`, `year`, `season` derived from `Date`
  - `discount_group` from `discount_percentage` — No discount (0%), Low
    (1–10%), High (11–20%)
  - Group summaries of revenue and units by discount group, category,
    day, and holiday status

# RO1: Seasonality & Timing — Lead: Jake Evans {background-color="#23351F"}

## RO1: Plan

- **Objective:** How do sales revenue and units sold vary across
  seasons, months, and holiday periods?
- **IVs:** month, season (derived from `Date`), Day of the Week, Holiday
  Effect
- **DVs:** Sales Revenue, Units Sold
- **Wrangling:** extract month/season from `Date`; aggregate revenue and
  units by month; compare holiday vs. non-holiday periods
- **Visualization:** monthly revenue line chart; holiday comparison
  bars; day-of-week heatmap; seasonal summary table

## RO1: Monthly Revenue Trend

```{r}
#| label: ro1-trend
#| code-fold: true
#| fig-height: 4.0
retail |>
  group_by(year_month) |>
  summarise(total_revenue = sum(sales_revenue_usd, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(year_month, total_revenue)) +
  geom_line(color = "#23351F", linewidth = 1) +
  geom_point(color = "#23351F", size = 1.8) +
  scale_y_continuous(labels = label_dollar(scale_cut = cut_short_scale())) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(x = NULL, y = "Total Revenue (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## RO1: Revenue by Season — Holiday vs. Non-Holiday

```{r}
#| label: ro1-holiday
#| code-fold: true
#| fig-height: 4.0
retail |>
  group_by(season, holiday) |>
  summarise(avg_revenue = mean(sales_revenue_usd, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(season, avg_revenue, fill = holiday)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_dollar()) +
  scale_fill_manual(values = c("Holiday" = "#D9A441", "Non-holiday" = "#4E6E8E")) +
  labs(x = NULL, y = "Avg Revenue per Transaction", fill = NULL)
```

## RO1: Revenue by Day of Week & Month

```{r}
#| label: ro1-heatmap
#| code-fold: true
#| fig-height: 4.0
retail |>
  group_by(month, day) |>
  summarise(avg_revenue = mean(sales_revenue_usd, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(month, day, fill = avg_revenue)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#EEF2EA", high = "#23351F",
                      labels = label_dollar()) +
  labs(x = NULL, y = NULL, fill = "Avg Revenue")
```

## RO1: Seasonal Summary

```{r}
#| label: ro1-table
#| code-fold: true
retail |>
  group_by(season) |>
  summarise(
    transactions = n(),
    avg_revenue  = mean(sales_revenue_usd, na.rm = TRUE),
    avg_units    = mean(units_sold, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(avg_revenue = dollar(avg_revenue, accuracy = 0.01),
         avg_units   = round(avg_units, 2)) |>
  kable(col.names = c("Season", "Transactions", "Avg Revenue", "Avg Units"))
```

# Research Objective Plans {background-color="#23351F"}

# RO2: Discounts & Units Sold — Lead: Aldo Naranjo {background-color="#23351F"}

## RO2: Plan

- **Objective:** Analyze whether higher discount percentages are
  associated with higher units sold
- **IV:** Discount Percentage
- **DV:** Units Sold
- **Controls:** Product Category, Store Location, Day of Week, Holiday
  Effect
- **Wrangling:** use existing `discount_group` variable (No / Low /
  High); summarize units sold by group
- **Visualizations:** scatterplot with trend line, boxplot by discount
  group, bar chart of average units sold

::: notes
Research Objective 2 examines the relationship between discount
percentage and units sold.

The independent variable is discount percentage, and the dependent
variable is units sold. Product category, store location, day of week,
and holiday effect are included as control variables since they may also
affect how many units are sold in a transaction.

For data preparation, we reuse the discount group variable already
created for the shared setup, grouping transactions into no discount,
low discount, and high discount categories. We then summarize units sold
within each group.

The planned visualizations include a scatterplot to see the overall
relationship, a boxplot to compare the distribution of units sold across
discount groups, and a bar chart summarizing average units sold per
group.
:::

## RO2: Discount Percentage vs. Units Sold

```{r}
#| label: ro2-scatter
#| code-fold: true
#| fig-height: 4.0

retail |>
  ggplot(aes(discount_percentage, units_sold)) +
  geom_jitter(alpha = 0.15,
              width = 0.3,
              height = 0,
              color = "#23351F") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "#D9A441",
    linewidth = 1
  ) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    x = "Discount Percentage",
    y = "Units Sold"
  )
```

::: notes
This scatterplot shows the overall relationship between discount
percentage and units sold across all transactions.

The x-axis represents the discount percentage applied to a transaction,
and the y-axis represents the number of units sold in that transaction.
Each point is one retail sale, and the fitted trend line summarizes the
general pattern across thousands of observations.

An upward-sloping trend line would suggest that larger discounts are
associated with more units sold. A flat or downward-sloping line would
suggest discounts have little effect, or even a negative one, on unit
volume.
:::

## RO2: Units Sold by Discount Group

```{r}
#| label: ro2-boxplot
#| code-fold: true
#| fig-height: 4.0

retail |>
  ggplot(aes(discount_group, units_sold, fill = discount_group)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c(
    "No discount" = "#4E6E8E",
    "Low (1–10%)" = "#D9A441",
    "High (11–20%)" = "#23351F"
  )) +
  labs(
    x = NULL,
    y = "Units Sold"
  )
```

::: notes
This boxplot compares the distribution of units sold across the three
discount groups: no discount, low discount, and high discount.

Rather than just comparing averages, the boxplot shows the spread and
median of units sold within each group, which helps identify whether
higher discount levels shift the whole distribution upward or simply
increase variability.
:::

## RO2: Average Units Sold Summary

```{r}
#| label: ro2-table
#| code-fold: true

retail |>
  group_by(discount_group) |>
  summarise(
    transactions = n(),
    avg_units = mean(units_sold, na.rm = TRUE),
    total_units = sum(units_sold, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(avg_units = round(avg_units, 2)) |>
  kable(
    col.names = c(
      "Discount Group",
      "Transactions",
      "Avg Units Sold",
      "Total Units Sold"
    )
  )
```

::: notes
This summary table provides the numerical results supporting the
previous charts.

For each discount group, the table reports the number of transactions,
the average units sold per transaction, and the total units sold. The
average units sold column lets us directly compare typical order sizes
across discount levels, while total units sold shows overall volume
contributed by each group.

Together with the scatterplot and boxplot, this table helps determine
whether higher discount percentages are actually associated with
meaningfully higher unit sales, which directly addresses Research
Objective 2.
:::

# RO3: Discounts & Revenue — Lead: Luis Berumen

## RO3: Plan

- **Objective:** Examine the relationship between discount percentage
  and sales revenue.
- **IV:** Discount Percentage
- **DV:** Sales Revenue (USD)
- **Controls:** Units Sold, Product Category, Marketing Spend
- **Wrangling:** Create discount groups; calculate transaction count,
  average revenue, and total revenue.
- **Visualizations:** Scatterplot, grouped bar chart, and revenue
  summary table.

::: notes
Research Objective 3 examines the relationship between discount
percentage and sales revenue.

The independent variable is discount percentage, and the dependent
variable is sales revenue. Units sold, product category, and marketing
spend are included as control variables because they may also influence
revenue.

For data preparation, transactions are grouped into no discount, low
discount, and high discount categories. We then calculate the number of
transactions, average revenue, and total revenue for each group.

The planned visualizations include a scatterplot to examine the overall
relationship, a grouped bar chart to compare the relationship across
product categories, and a summary table to provide the numerical
results.
:::

## RO3: Discount Percentage vs. Sales Revenue

```{r}
#| label: ro3-scatter
#| code-fold: true
#| fig-height: 4.0

retail |>
  ggplot(aes(discount_percentage, sales_revenue_usd)) +
  geom_jitter(alpha = 0.15,
              width = 0.3,
              height = 0,
              color = "#23351F") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "#D9A441",
    linewidth = 1
  ) +
  scale_y_continuous(labels = label_dollar()) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    x = "Discount Percentage",
    y = "Sales Revenue (USD)"
  )
```

::: notes
This scatterplot provides the broadest view of the relationship between
discount percentage and sales revenue.

The x-axis represents the discount percentage applied to each
transaction, while the y-axis represents the sales revenue generated
from that transaction. Each point corresponds to one retail sale.

Because the dataset contains thousands of observations, the points show
the overall distribution of the data, while the fitted trend line
summarizes the general relationship.

If the trend line slopes upward, it suggests that higher discount
percentages tend to be associated with higher sales revenue. If it
slopes downward, larger discounts may be associated with lower revenue.
A relatively flat trend line would suggest little overall relationship
between discount percentage and sales revenue.

This chart helps us evaluate both the direction and strength of the
relationship across the entire dataset.
:::

## RO3: Revenue by Product Category and Discount Group

```{r}
#| label: ro3-category
#| code-fold: true
#| fig-height: 4.0

retail |>
  group_by(product_category, discount_group) |>
  summarise(
    avg_revenue = mean(sales_revenue_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(product_category, avg_revenue, fill = discount_group)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_dollar()) +
  scale_fill_manual(values = c(
    "No discount" = "#4E6E8E",
    "Low (1–10%)" = "#D9A441",
    "High (11–20%)" = "#23351F"
  )) +
  labs(
    x = NULL,
    y = "Average Revenue (USD)",
    fill = NULL
  )
```

::: notes
This grouped bar chart explores whether the relationship between
discount percentage and sales revenue is consistent across different
product categories.

The x-axis displays each product category, while the colored bars
represent the three discount groups. The height of each bar represents
the average sales revenue for that combination of product category and
discount level.

By comparing the bars within the same category, we can evaluate whether
higher discount percentages are associated with changes in average
revenue for that specific product type. Comparing across categories
helps determine whether the relationship differs between Electronics,
Furniture, Clothing, and Groceries.

Including product category as a control variable provides additional
context and helps us determine whether the relationship between
discounts and revenue is consistent across different types of products.
:::

## RO3: Revenue Summary

```{r}
#| label: ro3-table
#| code-fold: true

retail |>
  group_by(discount_group) |>
  summarise(
    transactions = n(),
    avg_revenue = mean(sales_revenue_usd, na.rm = TRUE),
    total_revenue = sum(sales_revenue_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    avg_revenue = dollar(avg_revenue),
    total_revenue = dollar(total_revenue)
  ) |>
  kable(
    col.names = c(
      "Discount Group",
      "Transactions",
      "Avg Revenue",
      "Total Revenue"
    )
  )
```

::: notes
This summary table provides the numerical statistics that support the
previous visualizations.

For each discount group, the table reports the number of transactions,
the average revenue per transaction, and the total revenue generated.

The transaction count helps us understand the size of each discount
group, while the average revenue allows us to compare typical sales
across discount levels. Total revenue provides a broader view of overall
performance.

Together, these summary statistics complement the charts and help us
evaluate the relationship between discount percentage and sales revenue
from both a transaction-level and an overall business perspective.
:::

# RO4: Category Differences — Lead: Tim Arana {background-color="#23351F"}

## RO4: Plan

- **Objective:** Compare whether discount effects differ across product
  categories
- **IV:** Discount Percentage / Discount Group
- **DVs:** Units Sold and Sales Revenue
- **Moderator:** Product Category
- **Wrangling:** group records by product category and discount group;
  compute average units sold and average revenue
- **Visualization:** grouped bar charts and category-level comparison
  table

## RO4: Average Units Sold by Category and Discount Group

```{r}
#| label: ro4-units-category
#| code-fold: true
#| fig-height: 4.0

retail |>
  group_by(product_category, discount_group) |>
  summarise(
    avg_units = mean(units_sold, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(product_category, avg_units, fill = discount_group)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(
    "No discount" = "#4E6E8E",
    "Low (1–10%)" = "#D9A441",
    "High (11–20%)" = "#23351F"
  )) +
  labs(
    x = NULL,
    y = "Average Units Sold",
    fill = NULL
  )
```

## RO4: Average Revenue by Category and Discount Group

```{r}
#| label: ro4-revenue-category
#| code-fold: true
#| fig-height: 4.0

retail |>
  group_by(product_category, discount_group) |>
  summarise(
    avg_revenue = mean(sales_revenue_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(product_category, avg_revenue, fill = discount_group)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_dollar()) +
  scale_fill_manual(values = c(
    "No discount" = "#4E6E8E",
    "Low (1–10%)" = "#D9A441",
    "High (11–20%)" = "#23351F"
  )) +
  labs(
    x = NULL,
    y = "Average Revenue (USD)",
    fill = NULL
  )
```

## RO4: Category-Level Summary

```{r}
#| label: ro4-summary-table
#| code-fold: true

retail |>
  group_by(product_category, discount_group) |>
  summarise(
    transactions = n(),
    avg_units = mean(units_sold, na.rm = TRUE),
    avg_revenue = mean(sales_revenue_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    avg_units = round(avg_units, 2),
    avg_revenue = dollar(avg_revenue)
  ) |>
  kable(
    col.names = c(
      "Product Category",
      "Discount Group",
      "Transactions",
      "Avg Units",
      "Avg Revenue"
    )
  )
```

::: notes
RO4 examines whether discounts work the same way across product
categories.

The main independent variable is discount group, while the dependent
variables are average units sold and average sales revenue. Product
category acts as a moderator because discounts may affect categories
differently.

The first chart compares average units sold by discount group within
each category. The second chart compares average revenue using the same
structure. The summary table provides the supporting numbers.

This helps the retail audience avoid treating all products the same. If
some categories respond better to discounts than others, the company can
prioritize promotions for the categories where discounts create stronger
sales or revenue outcomes.
:::

# RO5: Marketing Spend Interaction — Lead: Mickyas Shawel {background-color="#23351F"}

## RO5: Plan

- **Objective:** Analyze whether marketing spend strengthens the impact
  of discounts on sales performance
- **IVs:** Discount Percentage and Marketing Spend (USD)
- **DVs:** Sales Revenue and Units Sold
- **Interaction:** Discount Percentage × Marketing Spend
- **Wrangling:** create marketing spend groups; summarize revenue and
  units sold by marketing spend group and discount group
- **Visualizations:** scatterplot with trend line, grouped bar chart,
  interaction plot, and summary table

::: notes
Research Objective 5 examines whether marketing spend strengthens the
relationship between discounts and retail sales performance.

The independent variables are discount percentage and marketing spend,
while the dependent variables are sales revenue and units sold. The main
idea is that discounts may perform better when they are supported by
higher marketing spend, because customers are more likely to become
aware of the promotion.

For data preparation, marketing spend is divided into low, medium, and
high groups. These groups are then compared across the existing discount
groups: no discount, low discount, and high discount.

The planned visualizations include a scatterplot showing the
relationship between marketing spend and sales revenue, a grouped bar
chart comparing average revenue by discount and marketing spend group,
and an interaction plot showing whether the discount effect changes as
marketing spend increases.
:::

## RO5: Marketing Spend vs. Sales Revenue

```{r}
#| label: ro5-scatter
#| code-fold: true
#| fig-height: 4.0

retail |>
  ggplot(aes(marketing_spend_usd, sales_revenue_usd)) +
  geom_jitter(
    alpha = 0.15,
    width = 0.3,
    height = 0,
    color = "#23351F"
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "#D9A441",
    linewidth = 1
  ) +
  scale_x_continuous(labels = label_dollar()) +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    x = "Marketing Spend (USD)",
    y = "Sales Revenue (USD)"
  )
```

::: notes
This scatterplot shows the overall relationship between marketing spend
and sales revenue.

The x-axis represents marketing spend, while the y-axis represents sales
revenue. Each point represents one retail sales record. The fitted trend
line summarizes whether higher marketing spend is generally associated
with higher sales revenue.

If the trend line slopes upward, it suggests that higher marketing spend
is associated with higher revenue. If the line is flat, it suggests
marketing spend may not have a strong relationship with revenue by
itself.
:::

## RO5: Average Revenue by Discount Group and Marketing Spend Group

```{r}
#| label: ro5-revenue-bar
#| code-fold: true
#| fig-height: 4.0

retail |>
  group_by(discount_group, marketing_spend_group) |>
  summarise(
    avg_revenue = mean(sales_revenue_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(discount_group, avg_revenue, fill = marketing_spend_group)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_dollar()) +
  scale_fill_manual(values = c(
    "Low marketing spend" = "#4E6E8E",
    "Medium marketing spend" = "#D9A441",
    "High marketing spend" = "#23351F"
  )) +
  labs(
    x = NULL,
    y = "Average Revenue (USD)",
    fill = NULL
  )
```

::: notes
This grouped bar chart compares average sales revenue across discount
groups and marketing spend groups.

The x-axis shows discount group, while each colored bar represents a
marketing spend group. The height of each bar represents average sales
revenue.

This chart helps show whether high marketing spend produces stronger
revenue outcomes within each discount level. If the high marketing spend
bars are consistently higher, it may suggest that marketing support
improves discount performance.
:::

## RO5: Discount and Marketing Spend Interaction

```{r}
#| label: ro5-interaction
#| code-fold: true
#| fig-height: 4.0

retail |>
  group_by(discount_group, marketing_spend_group) |>
  summarise(
    avg_revenue = mean(sales_revenue_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(
    discount_group,
    avg_revenue,
    group = marketing_spend_group,
    color = marketing_spend_group
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = label_dollar()) +
  scale_color_manual(values = c(
    "Low marketing spend" = "#4E6E8E",
    "Medium marketing spend" = "#D9A441",
    "High marketing spend" = "#23351F"
  )) +
  labs(
    x = NULL,
    y = "Average Revenue (USD)",
    color = NULL
  )
```

::: notes
This interaction plot is the main visual for Research Objective 5.

The chart shows whether the relationship between discount group and
average revenue changes depending on the level of marketing spend. Each
line represents a different marketing spend group.

If the lines are not parallel, that suggests an interaction effect. For
example, if the high marketing spend line rises more sharply across
discount groups than the low marketing spend line, it would suggest that
discounts are more effective when supported by higher marketing spend.
:::

## RO5: Marketing Spend Interaction Summary

```{r}
#| label: ro5-summary-table
#| code-fold: true

retail |>
  group_by(discount_group, marketing_spend_group) |>
  summarise(
    transactions = n(),
    avg_marketing_spend = mean(marketing_spend_usd, na.rm = TRUE),
    avg_units = mean(units_sold, na.rm = TRUE),
    avg_revenue = mean(sales_revenue_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    avg_marketing_spend = dollar(avg_marketing_spend),
    avg_units = round(avg_units, 2),
    avg_revenue = dollar(avg_revenue)
  ) |>
  kable(
    col.names = c(
      "Discount Group",
      "Marketing Spend Group",
      "Transactions",
      "Avg Marketing Spend",
      "Avg Units Sold",
      "Avg Revenue"
    )
  )
```

::: notes
This summary table provides the numerical results behind the RO5 charts.

For each combination of discount group and marketing spend group, the
table shows the number of transactions, average marketing spend, average
units sold, and average sales revenue.

This table helps support the visual interpretation by showing whether
higher marketing spend is associated with stronger sales outcomes within
each discount level.
:::

# Conclusion {background-color="#23351F"}

## Planned Analysis & Expected Contribution

- **Procedure:** wrangle once (shared setup), then group summaries and
  visualizations per RO; optional model direction — regression of
  units/revenue on discount, marketing spend, and controls
- **What the audience gains:**
  - When discounts are useful vs. when they erode revenue efficiency
  - Which product categories respond to discounts
  - Whether marketing spend amplifies discount effects
  - How seasonal and holiday timing shapes discount performance
- **Deliverable becomes actionable:** discount levels, category
  priorities, marketing pairing, and a seasonal promotion calendar

## Conclusion & Next Steps

- **Problem:** discounts drive volume but may erode revenue — retailers
  need evidence, not habit
- **Approach:** five ROs covering timing, volume, revenue, category, and
  marketing interaction on 30,000 retail transactions
- **Next steps toward Ch4 (Results) and Ch5 (Recommendations):** each
  lead finalizes summaries, visualizations, and interpretation for their
  RO

## Thank You

**Questions?**

Group 1 — IBM 6540 \| Summer 2026

Repo: <https://github.com/mjshawell/IBM-6540-Group-Project>
