---
title: "Analyzing the Effect of Discounts on Retail Sales Volume and Revenue Using Seasonal Retail Data"
subtitle: "Project Proposal — Group 1"
author: "Jake Evans · Mickyas Shawel · Aldo Naranjo · Luis Berumen · Tim Arana"
institute: "IBM 6540 — Applied Machine Learning for Marketing | Cal Poly Pomona"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    embed-resources: true
    footer: "IBM 6540 — Group 1 Project Proposal | Summer 2026"
    toc: false
execute:
  warning: false
  message: false
---

```{r}
#| label: global-setup
#| include: false
library(tidyverse)
library(janitor)
library(lubridate)
library(scales)
library(knitr)
theme_set(theme_minimal(base_size = 14))

csv_files <- list.files(".", pattern = "\\.csv$", recursive = TRUE, full.names = TRUE)
data_path <- csv_files[grepl("retail|sales", csv_files, ignore.case = TRUE)][1]
if (is.na(data_path)) data_path <- csv_files[1]

retail <- read_csv(data_path) |>
  clean_names() |>
  mutate(
    date       = as.Date(date),
    month      = month(date, label = TRUE, abbr = TRUE),
    year       = year(date),
    year_month = floor_date(date, "month"),
    season = case_when(
      month(date) %in% c(12, 1, 2)  ~ "Winter",
      month(date) %in% c(3, 4, 5)   ~ "Spring",
      month(date) %in% c(6, 7, 8)   ~ "Summer",
      TRUE                          ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    holiday = if_else(as.logical(holiday_effect), "Holiday", "Non-holiday"),
    day = factor(day_of_the_week,
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday",
                 "Friday", "Saturday", "Sunday")),
    discount_group = case_when(
      discount_percentage == 0  ~ "No discount",
      discount_percentage <= 10 ~ "Low (1–10%)",
      TRUE                      ~ "High (11–20%)"
    ),
    discount_group = factor(discount_group,
      levels = c("No discount", "Low (1–10%)", "High (11–20%)"))
  )
```

## Project Overview

**Main research question:** How does discount percentage affect retail sales volume and revenue when accounting for product category, marketing spend, holidays, and seasonal trends?

| Member         | RO Lead |
|----------------|---------|
| Jake Evans     | RO 1    |
| Mickyas Shawel | RO \_\_ |
| Aldo Naranjo   | RO \_\_ |
| Luis Berumen   | RO 3    |
| Tim Arana      | RO \_\_ |

::: aside
Dataset: [Retail Sales Data with Seasonal Trends & Marketing (Kaggle)](https://www.kaggle.com/datasets/abdullah0a/retail-sales-data-with-seasonal-trends-and-marketing) · Repo: <https://github.com/mjshawell/IBM-6540-Group-Project>
:::

# Ch1: Introduction {background-color="#23351F"}

## Problem Definition

- **Audience:** retail managers, ecommerce teams, marketing analysts, and promotion planners who need to evaluate discount strategy
- Retailers frequently use discounts to attract customers, increase demand, and clear inventory
- **The tradeoff:** discounts may increase units sold while reducing revenue per unit
- Without data analysis, retailers may overuse discounts, underuse them, or apply the wrong discount levels to the wrong products or time periods

## Research Objectives

- **RO1 — Seasonality & Timing:** How do sales revenue and units sold vary across seasons, months, and holiday periods?
- **RO2 — Discounts & Volume:** Are higher discount percentages associated with higher units sold?
- **RO3 — Discounts & Revenue:** Do discounts increase sales revenue?
- **RO4 — Category Differences:** Does the effect of discounts differ across product categories?
- **RO5 — Marketing Interaction:** Does marketing spend strengthen the impact of discounts?

::: aside
Each member leads one RO — confirm assignments on slide 2.
:::

## Importance of the Topic

- Helps retailers decide when discounts are useful versus when they reduce revenue efficiency
- Supports product-level promotional planning by identifying categories that respond well to discounts
- Improves marketing strategy by showing whether discounts perform better with higher marketing spend
- Supports seasonal planning by comparing holiday and non-holiday performance
- Creates a data-driven foundation for pricing, promotion, and revenue optimization decisions

## Background & Hypothesis Directions

| RO | Hypothesis Direction |
|------------------|------------------------------------------------------|
| RO1 | Revenue and units will vary by season, with holiday periods outperforming regular periods |
| RO2 | Higher discount percentages will be associated with higher units sold |
| RO3 | Discounts may lift revenue up to a point, but higher discounts may not produce proportional gains |
| RO4 | The discount–sales relationship will vary by product category |
| RO5 | Discounts paired with higher marketing spend will show stronger sales outcomes |

# Ch3: Methods {background-color="#23351F"}

## Dataset Overview & Key Variables

- **Source:** Kaggle secondary dataset (synthetic, designed for analytics practice) — 30,000 transaction-level records, 2022–2024
- **DVs:** Sales Revenue (USD), Units Sold
- **Primary IV:** Discount Percentage (0–20%)
- **Additional predictors:** Marketing Spend (USD), Product Category, Store Location, Day of the Week, Holiday Effect, Date

## Population, Sampling & Data Collection

- **Population represented:** transaction-level retail sales records across stores, locations, and product categories
- The dataset is synthetic; it stands in for multi-location retail chains that run discount and marketing programs
- **Sampling:** secondary data collection — records are already collected and will be prepared for analysis in R

## Data Wrangling Plan

- **From records to data:** import CSV with `readr::read_csv()`; each row = one retail sales record
- **Tidying:** `janitor::clean_names()`; check missing values, duplicates, and outliers; convert `Date` to date type; standardize categorical variables (category, location, day, holiday)
- **Transforming:**
  - `month`, `year`, `season` derived from `Date`
  - `discount_group` from `discount_percentage` — No discount (0%), Low (1–10%), High (11–20%)
  - Group summaries of revenue and units by discount group, category, day, and holiday status

# RO1: Seasonality & Timing — Lead: Jake Evans {background-color="#23351F"}

## RO1: Plan

- **Objective:** How do sales revenue and units sold vary across seasons, months, and holiday periods?
- **IVs:** month, season (derived from `Date`), Day of the Week, Holiday Effect
- **DVs:** Sales Revenue, Units Sold
- **Wrangling:** extract month/season from `Date`; aggregate revenue and units by month; compare holiday vs. non-holiday periods
- **Visualization:** monthly revenue line chart; holiday comparison bars; day-of-week heatmap; seasonal summary table

## RO1: Monthly Revenue Trend

```{r}
#| label: ro1-trend
#| code-fold: true
#| fig-height: 4.5
retail |>
  group_by(year_month) |>
  summarise(total_revenue = sum(sales_revenue_usd, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(year_month, total_revenue)) +
  geom_line(color = "#23351F", linewidth = 1) +
  geom_point(color = "#23351F", size = 1.8) +
  scale_y_continuous(labels = label_dollar(scale_cut = cut_short_scale())) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(x = NULL, y = "Total Revenue (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## RO1: Revenue by Season — Holiday vs. Non-Holiday

```{r}
#| label: ro1-holiday
#| code-fold: true
#| fig-height: 4.5
retail |>
  group_by(season, holiday) |>
  summarise(avg_revenue = mean(sales_revenue_usd, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(season, avg_revenue, fill = holiday)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_dollar()) +
  scale_fill_manual(values = c("Holiday" = "#D9A441", "Non-holiday" = "#4E6E8E")) +
  labs(x = NULL, y = "Avg Revenue per Transaction", fill = NULL)
```

## RO1: Revenue by Day of Week & Month

```{r}
#| label: ro1-heatmap
#| code-fold: true
#| fig-height: 4.5
retail |>
  group_by(month, day) |>
  summarise(avg_revenue = mean(sales_revenue_usd, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(month, day, fill = avg_revenue)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#EEF2EA", high = "#23351F",
                      labels = label_dollar()) +
  labs(x = NULL, y = NULL, fill = "Avg Revenue")
```

## RO1: Seasonal Summary

```{r}
#| label: ro1-table
#| code-fold: true
retail |>
  group_by(season) |>
  summarise(
    transactions = n(),
    avg_revenue  = mean(sales_revenue_usd, na.rm = TRUE),
    avg_units    = mean(units_sold, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(avg_revenue = dollar(avg_revenue, accuracy = 0.01),
         avg_units   = round(avg_units, 2)) |>
  kable(col.names = c("Season", "Transactions", "Avg Revenue", "Avg Units"))
```

# Research Objective Plans {background-color="#23351F"}

# RO2: Discounts & Units Sold — Lead: Aldo Naranjo {background-color="#23351F"}

## RO2: Plan

- **Objective:** Analyze whether higher discount percentages are associated with higher units sold
- **IV:** Discount Percentage
- **DV:** Units Sold
- **Controls:** Product Category, Store Location, Day of Week, Holiday Effect
- **Wrangling:** use existing `discount_group` variable (No / Low / High); summarize units sold by group
- **Visualizations:** scatterplot with trend line, boxplot by discount group, bar chart of average units sold

::: {.notes}
Research Objective 2 examines the relationship between discount percentage and
units sold.

The independent variable is discount percentage, and the dependent variable is
units sold. Product category, store location, day of week, and holiday effect
are included as control variables since they may also affect how many units
are sold in a transaction.

For data preparation, we reuse the discount group variable already created for
the shared setup, grouping transactions into no discount, low discount, and
high discount categories. We then summarize units sold within each group.

The planned visualizations include a scatterplot to see the overall
relationship, a boxplot to compare the distribution of units sold across
discount groups, and a bar chart summarizing average units sold per group.
:::

## RO2: Discount Percentage vs. Units Sold

```{r}
#| label: ro2-scatter
#| code-fold: true
#| fig-height: 4.5

retail |>
  ggplot(aes(discount_percentage, units_sold)) +
  geom_jitter(alpha = 0.15,
              width = 0.3,
              height = 0,
              color = "#23351F") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "#D9A441",
    linewidth = 1
  ) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    x = "Discount Percentage",
    y = "Units Sold"
  )
```

::: {.notes}
This scatterplot shows the overall relationship between discount percentage
and units sold across all transactions.

The x-axis represents the discount percentage applied to a transaction, and
the y-axis represents the number of units sold in that transaction. Each point
is one retail sale, and the fitted trend line summarizes the general pattern
across thousands of observations.

An upward-sloping trend line would suggest that larger discounts are
associated with more units sold. A flat or downward-sloping line would suggest
discounts have little effect, or even a negative one, on unit volume.
:::

## RO2: Units Sold by Discount Group

```{r}
#| label: ro2-boxplot
#| code-fold: true
#| fig-height: 4.5

retail |>
  ggplot(aes(discount_group, units_sold, fill = discount_group)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c(
    "No discount" = "#4E6E8E",
    "Low (1–10%)" = "#D9A441",
    "High (11–20%)" = "#23351F"
  )) +
  labs(
    x = NULL,
    y = "Units Sold"
  )
```

::: {.notes}
This boxplot compares the distribution of units sold across the three
discount groups: no discount, low discount, and high discount.

Rather than just comparing averages, the boxplot shows the spread and median
of units sold within each group, which helps identify whether higher discount
levels shift the whole distribution upward or simply increase variability.
:::

## RO2: Average Units Sold Summary

```{r}
#| label: ro2-table
#| code-fold: true

retail |>
  group_by(discount_group) |>
  summarise(
    transactions = n(),
    avg_units = mean(units_sold, na.rm = TRUE),
    total_units = sum(units_sold, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(avg_units = round(avg_units, 2)) |>
  kable(
    col.names = c(
      "Discount Group",
      "Transactions",
      "Avg Units Sold",
      "Total Units Sold"
    )
  )
```

::: {.notes}
This summary table provides the numerical results supporting the previous
charts.

For each discount group, the table reports the number of transactions, the
average units sold per transaction, and the total units sold. The average
units sold column lets us directly compare typical order sizes across
discount levels, while total units sold shows overall volume contributed by
each group.

Together with the scatterplot and boxplot, this table helps determine whether
higher discount percentages are actually associated with meaningfully higher
unit sales, which directly addresses Research Objective 2.
:::

# RO3: Discounts & Revenue — Lead: Luis Berumen

## RO3: Plan

- **Objective:** Examine the relationship between discount percentage and sales revenue.
- **IV:** Discount Percentage
- **DV:** Sales Revenue (USD)
- **Controls:** Units Sold, Product Category, Marketing Spend
- **Wrangling:** Create discount groups; calculate transaction count, average revenue, and total revenue.
- **Visualizations:** Scatterplot, grouped bar chart, and revenue summary table.

::: {.notes}
Research Objective 3 examines the relationship between discount percentage and
sales revenue.

The independent variable is discount percentage, and the dependent variable is
sales revenue. Units sold, product category, and marketing spend are included as
control variables because they may also influence revenue.

For data preparation, transactions are grouped into no discount, low discount,
and high discount categories. We then calculate the number of transactions,
average revenue, and total revenue for each group.

The planned visualizations include a scatterplot to examine the overall
relationship, a grouped bar chart to compare the relationship across product
categories, and a summary table to provide the numerical results.
:::

## RO3: Discount Percentage vs. Sales Revenue

```{r}
#| label: ro3-scatter
#| code-fold: true
#| fig-height: 4.5

retail |>
  ggplot(aes(discount_percentage, sales_revenue_usd)) +
  geom_jitter(alpha = 0.15,
              width = 0.3,
              height = 0,
              color = "#23351F") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "#D9A441",
    linewidth = 1
  ) +
  scale_y_continuous(labels = label_dollar()) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    x = "Discount Percentage",
    y = "Sales Revenue (USD)"
  )
```

::: {.notes}
This scatterplot provides the broadest view of the relationship between discount
percentage and sales revenue.

The x-axis represents the discount percentage applied to each transaction, while
the y-axis represents the sales revenue generated from that transaction. Each
point corresponds to one retail sale.

Because the dataset contains thousands of observations, the points show the
overall distribution of the data, while the fitted trend line summarizes the
general relationship.

If the trend line slopes upward, it suggests that higher discount percentages
tend to be associated with higher sales revenue. If it slopes downward, larger
discounts may be associated with lower revenue. A relatively flat trend line
would suggest little overall relationship between discount percentage and sales
revenue.

This chart helps us evaluate both the direction and strength of the relationship
across the entire dataset.
:::

## RO3: Revenue by Product Category and Discount Group

```{r}
#| label: ro3-category
#| code-fold: true
#| fig-height: 4.5

retail |>
  group_by(product_category, discount_group) |>
  summarise(
    avg_revenue = mean(sales_revenue_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(product_category, avg_revenue, fill = discount_group)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_dollar()) +
  scale_fill_manual(values = c(
    "No discount" = "#4E6E8E",
    "Low (1–10%)" = "#D9A441",
    "High (11–20%)" = "#23351F"
  )) +
  labs(
    x = NULL,
    y = "Average Revenue (USD)",
    fill = NULL
  )
```

::: {.notes}
This grouped bar chart explores whether the relationship between discount
percentage and sales revenue is consistent across different product categories.

The x-axis displays each product category, while the colored bars represent the
three discount groups. The height of each bar represents the average sales
revenue for that combination of product category and discount level.

By comparing the bars within the same category, we can evaluate whether higher
discount percentages are associated with changes in average revenue for that
specific product type. Comparing across categories helps determine whether the
relationship differs between Electronics, Furniture, Clothing, and Groceries.

Including product category as a control variable provides additional context and
helps us determine whether the relationship between discounts and revenue is
consistent across different types of products.
:::

## RO3: Revenue Summary

```{r}
#| label: ro3-table
#| code-fold: true

retail |>
  group_by(discount_group) |>
  summarise(
    transactions = n(),
    avg_revenue = mean(sales_revenue_usd, na.rm = TRUE),
    total_revenue = sum(sales_revenue_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    avg_revenue = dollar(avg_revenue),
    total_revenue = dollar(total_revenue)
  ) |>
  kable(
    col.names = c(
      "Discount Group",
      "Transactions",
      "Avg Revenue",
      "Total Revenue"
    )
  )
```

::: {.notes}
This summary table provides the numerical statistics that support the previous
visualizations.

For each discount group, the table reports the number of transactions, the
average revenue per transaction, and the total revenue generated.

The transaction count helps us understand the size of each discount group, while
the average revenue allows us to compare typical sales across discount levels.
Total revenue provides a broader view of overall performance.

Together, these summary statistics complement the charts and help us evaluate
the relationship between discount percentage and sales revenue from both a
transaction-level and an overall business perspective.
:::

## RO4: Category Differences — Lead: \_\_\_\_\_\_

- **Objective:** Compare the effect of discounts across product categories
- **IV:** Discount Percentage · **DV:** Units Sold / Sales Revenue
- **Moderator:** Product Category (Clothing, Electronics, Furniture, Groceries)
- **Wrangling plan:** group records by product category and discount group; compute average revenue and units sold
- **Visualization plan:** faceted bar charts by category; category-level comparison table; grouped bar chart

## RO5: Marketing Spend Interaction — Lead: \_\_\_\_\_\_

- **Objective:** Analyze whether marketing spend strengthens the impact of discounts
- **IVs:** Discount Percentage, Marketing Spend (USD) · **DV:** Sales Revenue / Units Sold
- **Possible interaction:** Discount Percentage × Marketing Spend
- **Wrangling plan:** create marketing spend groups; summarize outcomes by marketing group and discount group; optional interaction model
- **Visualization plan:** scatterplot of marketing spend vs. sales; grouped bar chart by discount group; interaction plot

# Conclusion {background-color="#23351F"}

## Planned Analysis & Expected Contribution

- **Procedure:** wrangle once (shared setup), then group summaries and visualizations per RO; optional model direction — regression of units/revenue on discount, marketing spend, and controls
- **What the audience gains:**
  - When discounts are useful vs. when they erode revenue efficiency
  - Which product categories respond to discounts
  - Whether marketing spend amplifies discount effects
  - How seasonal and holiday timing shapes discount performance
- **Deliverable becomes actionable:** discount levels, category priorities, marketing pairing, and a seasonal promotion calendar

## Conclusion & Next Steps

- **Problem:** discounts drive volume but may erode revenue — retailers need evidence, not habit
- **Approach:** five ROs covering timing, volume, revenue, category, and marketing interaction on 30,000 retail transactions
- **Next steps toward Ch4 (Results) and Ch5 (Recommendations):** each lead finalizes summaries, visualizations, and interpretation for their RO

## Thank You

**Questions?**

Group 1 — IBM 6540 \| Summer 2026

Repo: <https://github.com/mjshawell/IBM-6540-Group-Project>
