---
title: "The Effect of Discounts on Retail Sales"
subtitle: "A Machine Learning Analysis of Discount Strategy and Revenue Performance"
author: "Group 1 — Jake Evans, Mickyas Shawel, Aldo Naranjo, Luis Berumen, Tim Arana"
institute: "IBM 6540 | Applied Machine Learning for Marketing | Cal Poly Pomona"
date: today
date-format: "MMMM D, YYYY"
format:
  revealjs:
    theme: theme.scss
    slide-number: true
    toc: true
    toc-depth: 1
    toc-title: "Agenda"
    embed-resources: true
    width: 1600
    height: 900
    margin: 0.06
    minScale: 0.35
    maxScale: 1.4
    transition: fade
    transition-speed: fast
    footer: "IBM 6540 | Group 1 | Cal Poly Pomona"
execute:
  warning: false
  message: false
  echo: false
  freeze: true
---

```{=html}
<style>
/* Global slide compacting to prevent overflow */
.reveal {
  font-size: 24px;
  line-height: 1.18;
}

.reveal .slides {
  text-align: left;
}

.reveal h1 {
  font-size: 1.75em;
  line-height: 1.02;
  margin-bottom: 0.25em;
}

.reveal h2 {
  font-size: 1.25em;
  line-height: 1.05;
  margin-bottom: 0.30em;
}

.reveal h3 {
  font-size: 0.95em;
  line-height: 1.08;
  margin-bottom: 0.22em;
}

.reveal p,
.reveal li,
.reveal blockquote,
.reveal td,
.reveal th {
  font-size: 0.76em;
  line-height: 1.18;
}

.reveal ul,
.reveal ol {
  margin-top: 0.2em;
  margin-bottom: 0.2em;
}

.reveal li {
  margin-bottom: 0.12em;
}

.reveal blockquote {
  padding: 0.35em 0.55em;
  margin: 0.35em 0;
}

.reveal table {
  font-size: 0.70em;
  line-height: 1.05;
  margin-top: 0.2em;
}

.reveal .cell-output-display,
.reveal .quarto-figure,
.reveal img,
.reveal svg {
  max-height: 470px !important;
  max-width: 100% !important;
}

.reveal pre code {
  font-size: 0.58em;
  line-height: 1.05;
}

.reveal .footer,
.reveal .slide-number {
  font-size: 0.42em !important;
}

.reveal .columns {
  gap: 0.8em;
}

.reveal .callout {
  font-size: 0.78em;
  padding: 0.35em 0.55em;
  margin: 0.3em 0;
}

#title-slide .title {
  font-size: 1.55em;
  line-height: 1.02;
  margin-bottom: 0.25em;
}

#title-slide .subtitle {
  font-size: 0.72em;
  line-height: 1.05;
  margin-bottom: 0.35em;
}

#title-slide .author,
#title-slide .institute,
#title-slide .date {
  font-size: 0.58em;
  line-height: 1.08;
}

/* Manual compact classes for dense slides */
.compact-slide p,
.compact-slide li {
  font-size: 0.70em;
  line-height: 1.12;
}

.compact-slide h2 {
  font-size: 1.12em;
}
</style>
```

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(tidymodels)
library(janitor)
library(lubridate)
library(scales)
library(readxl)
library(gt)
library(patchwork)
library(vip)
library(rpart.plot)
tidymodels_prefer()
set.seed(617)

retail <- read_excel("Retail_sales.xlsx") |>
  clean_names() |>
  rename(
    revenue      = sales_revenue_usd,
    discount_pct = discount_percentage,
    mktg_spend   = marketing_spend_usd,
    units        = units_sold,
    category     = product_category,
    location     = store_location,
    day          = day_of_the_week,
    holiday      = holiday_effect
  ) |>
  mutate(
    date   = as.Date(date),
    season = case_when(
      month(date) %in% c(12, 1, 2) ~ "Winter",
      month(date) %in% c(3, 4, 5)  ~ "Spring",
      month(date) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                          ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    holiday = factor(if_else(as.logical(holiday), "Holiday", "Non-Holiday")),
    discount_group = case_when(
      discount_pct == 0  ~ "No Discount",
      discount_pct <= 10 ~ "Low (1-10%)",
      TRUE               ~ "High (11-20%)"
    ),
    discount_group = factor(
      discount_group,
      levels = c("No Discount", "Low (1-10%)", "High (11-20%)")
    ),
    category = factor(category),
    day      = factor(day),
    high_revenue = factor(
      if_else(revenue >= median(revenue, na.rm = TRUE), "high", "low"),
      levels = c("high", "low")
    )
  )
```

# Introduction

## Problem Definition

::::: columns
::: {.column width="55%"}
Retailers face a persistent tension between offering discounts to drive volume and protecting per-transaction revenue. Key questions include:

- Do discounts actually increase units sold?
- At what point do discounts reduce net revenue?
- Which product categories and seasons respond differently to promotional pricing?
- Can a machine learning model reliably predict revenue outcomes from discount strategy?
:::

::: {.column width="45%"}
```{r}
#| fig-height: 3.8
retail |>
  group_by(discount_group) |>
  summarise(
    avg_rev   = mean(revenue),
    avg_units = mean(units)
  ) |>
  ggplot(aes(discount_group, avg_rev, fill = discount_group)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#4E6E8E", "#FFB81C", "#00693D")) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Avg Revenue by Discount Level",
       x = NULL, y = "Avg Revenue (USD)") +
  theme_minimal(base_size = 10)
```
:::
:::::

## Research Objectives

::::: columns
::: {.column width="50%"}

**Five research objectives guide our analysis:**

| RO  | Member          | Focus                         |
|-----|-----------------|-------------------------------|
| RO1 | Jake Evans      | Seasonality & Holiday Effects |
| RO2 | Aldo Naranjo    | Discount % vs. Units Sold     |
| RO3 | Luis Berumen    | Discounts & Sales Revenue     |
| RO4 | Tim Arana       | Sales Revenue Prediction      |
| RO5 | Mickyas Shawel  | Marketing Spend Interaction   |

:::

::: {.column width="50%"}

**Our overarching analytics objective:**

> Build supervised machine learning models that predict retail sales performance from discount strategy, marketing spend, product category, and seasonal features using 30,000 transaction-level records from 2022 to 2024.

:::
:::::

## Why This Research Matters 

- **For retailers:** Discounting is one of the most common promotional tools, but the margin impact is often unclear. Understanding the revenue consequences of different discount levels by category and season allows more precise promotional planning.

- **For marketing analysts:** Machine learning models that predict whether a transaction will generate above-median revenue give analysts a decision tool that goes beyond aggregate reporting.

- **For the field:** Most existing research treats discounts in isolation. This project examines discount effects alongside marketing spend, seasonality, holiday effects, and product category simultaneously.

# Population & Sample

## Population and Sample 

:::::: columns
:::: {.column width="50%"}
**Population of interest:** Retail transactions across product categories in a multi-store, multi-region retail environment.

**Sample:** 30,000 transaction-level records from a Kaggle synthetic retail dataset designed to reflect realistic retail patterns (2022–2024).

::: callout-note
The group is using synthetic data for this analysis. Real transaction data was not available for this project. Findings should be interpreted in the context of the synthetic data generating process.
:::
::::

::: {.column width="50%"}
```{r}
#| fig-height: 3.8
retail |>
  count(category) |>
  ggplot(aes(reorder(category, n), n, fill = category)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c("#00693D","#FFB81C","#4E6E8E","#8B3A3A")) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Transactions by Category", x = NULL, y = "Count") +
  theme_minimal(base_size = 10)
```
:::
::::::

## Sample Characteristics

```{r}
retail |>
  summarise(
    `Total Transactions`     = scales::comma(n()),
    `Date Range`             = paste(min(date), "to", max(date)),
    `Avg Revenue`            = dollar(mean(revenue)),
    `Median Revenue`         = dollar(median(revenue)),
    `Avg Discount %`         = paste0(round(mean(discount_pct), 1), "%"),
    `Avg Units Sold`         = as.character(round(mean(units), 1)),
    `Avg Marketing Spend`    = dollar(mean(mktg_spend)),
    `Holiday Transactions %` = paste0(round(mean(holiday == "Holiday") * 100, 1), "%")
  ) |>
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") |>
  gt() |>
  tab_header(title = "Key Dataset Summary Statistics") |>
  cols_align(align = "left", columns = Metric) |>
  cols_align(align = "right", columns = Value) |>
  tab_options(table.font.size = px(12))
```

------------------------------------------------------------------------

# RO1 — Jake Evans

## RO1: Machine Learning Methods

- **Objective:** Predict Sales Revenue from seasonal and timing features — season, month, day of week, and holiday status — alongside product category, discount group, and marketing spend group
- **Regression-Focused Method:** Elastic Net regression (tuned penalty and mixture) as the interpretable baseline
- **Ensemble Method:** Random Forest regression (primary model) to capture non-linear seasonal interactions
- **Evaluation Metric:** RMSE and R-squared (regression)

::: notes
RO1 predicts sales revenue as a continuous outcome rather than a high/low class, which lets us quantify exactly how much seasonal timing moves revenue in dollar terms.

Elastic Net is used as an interpretable, regularized baseline. Random Forest is the primary model because seasonal effects often interact with category and discount group in non-linear ways that a linear model can't capture.
:::

## RO1: Model Fitting

```{r}
#| label: ro1-ml-split
#| echo: true
#| code-fold: true

set.seed(617)

ro1_data <- retail |>
  select(revenue, season, day, holiday, category, discount_group, mktg_spend)

split <- initial_split(ro1_data, prop = 0.80, strata = revenue)
train <- training(split)
test  <- testing(split)

rec <- recipe(revenue ~ ., data = train) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

en_spec <- linear_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("regression")

en_wf <- workflow() |> add_recipe(rec) |> add_model(en_spec)

folds   <- vfold_cv(train, v = 5, strata = revenue)
en_grid <- grid_regular(penalty(), mixture(), levels = 5)

en_tuned <- tune_grid(en_wf, resamples = folds, grid = en_grid,
                       metrics = metric_set(rmse, rsq))
best_en  <- select_best(en_tuned, metric = "rmse")
en_final <- finalize_workflow(en_wf, best_en) |> fit(train)

rf_spec <- rand_forest(trees = 300, mtry = 4, min_n = 10) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("regression")

rf_wf  <- workflow() |> add_recipe(rec) |> add_model(rf_spec)
rf_fit <- fit(rf_wf, data = train)
```

## RO1: Model Evaluation

::::: columns
::: {.column width="50%"}
```{r}
#| label: ro1-ml-metrics
#| code-fold: true

en_preds <- augment(en_final, new_data = test)
rf_preds <- augment(rf_fit, new_data = test)

en_metrics <- en_preds |>
  metrics(truth = revenue, estimate = .pred) |>
  mutate(model = "Elastic Net")

rf_metrics <- rf_preds |>
  metrics(truth = revenue, estimate = .pred) |>
  mutate(model = "Random Forest")

bind_rows(en_metrics, rf_metrics) |>
  filter(.metric %in% c("rmse", "rsq")) |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  mutate(rmse = dollar(rmse, accuracy = 0.01),
         rsq  = round(rsq, 3)) |>
  knitr::kable(col.names = c("Model", "RMSE", "R-squared"))
```
:::

::: {.column width="50%"}
```{r}
#| label: ro1-ml-scatter
#| code-fold: true
#| fig-height: 4.0

rf_preds |>
  ggplot(aes(revenue, .pred)) +
  geom_point(alpha = 0.15, color = "#23351F") +
  geom_abline(linetype = "dashed", color = "#D9A441", linewidth = 1) +
  scale_x_continuous(labels = label_dollar()) +
  scale_y_continuous(labels = label_dollar()) +
  labs(x = "Actual Revenue", y = "Predicted Revenue")
```
:::
:::::

::: notes
The metrics table compares Elastic Net and Random Forest on RMSE (average dollar error) and R-squared (share of revenue variation explained). The scatterplot shows predicted vs. actual revenue for the Random Forest — points closer to the dashed line mean more accurate predictions.
:::

## RO1: Variable Importance

::::: columns
::: {.column width="55%"}
```{r}
#| label: ro1-ml-vip
#| fig-height: 5.0
#| code-fold: true

rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 10,
      aesthetics = list(fill = "#23351F", color = "white", alpha = 0.85)) +
  labs(title = "Variable Importance — Random Forest (RO1)") +
  theme_minimal(base_size = 13)
```
:::

::: {.column width="45%"}
**Key findings from VIP:**

- Product category and discount group tend to dominate over pure seasonal timing
- Season and holiday status still contribute meaningfully, confirming timing effects exist but are secondary to category/discount
- Day of week is typically the weakest predictor
:::
:::::

::: notes
This chart ranks which features the Random Forest relied on most to predict revenue. If category or discount group outrank season, it means seasonal timing shifts revenue, but category and promotion strategy still explain more of the variation overall.
:::

## RO1: Summary Findings

::::: columns
::: {.column width="60%"}
1.  Revenue varies measurably by season and holiday status, but the effect is smaller than category or discount-driven variation
2.  Random Forest outperforms Elastic Net on RMSE, indicating seasonal effects interact non-linearly with category and discount group
3.  Holiday periods show a consistent revenue lift over non-holiday periods within the same season
4.  Timing alone is not sufficient for revenue forecasting — it should be combined with category and promotional context
:::

::: {.column width="40%"}
**Business recommendation:**

> Use seasonal and holiday timing as a secondary lever layered on top of category- and discount-driven strategy, not as a standalone driver of revenue planning.
:::
:::::

------------------------------------------------------------------------

# RO2 — Aldo Naranjo

## RO2: Do Discounts Drive Unit Sales? 

**Analytics Objective:** Predict whether a retail transaction generates above- or below-median revenue (`high` vs. `low`) using discount percentage, marketing spend, units sold, product category, season, and holiday status.

::::: columns
::: {.column width="55%"}
**ML Methods Used:**

- Lasso Logistic Regression (baseline)
- Tuned Decision Tree
- **Random Forest (primary)**
- XGBoost

**Evaluation Metric:** ROC AUC (classification)
:::

::: {.column width="45%"}
```{r}
#| fig-height: 3.6
retail |>
  ggplot(aes(x = discount_pct, y = units, color = high_revenue)) +
  geom_jitter(alpha = 0.08, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("high" = "#00693D", "low" = "#8B3A3A")) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(title = "Discount % vs. Units Sold",
       subtitle = "Colored by revenue class",
       x = "Discount %", y = "Units Sold", color = "Revenue") +
  theme_minimal(base_size = 10)
```
:::
:::::

## RO2: Data Wrangling & EDA 

```{r}
#| fig-height: 3.8
p1 <- retail |>
  group_by(category, discount_group) |>
  summarise(avg_units = mean(units), .groups = "drop") |>
  ggplot(aes(category, avg_units, fill = discount_group)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#4E6E8E","#FFB81C","#00693D")) +
  labs(title = "Avg Units by Category & Discount",
       x = NULL, y = "Avg Units", fill = "Discount") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "top")

p2 <- retail |>
  group_by(season, high_revenue) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(season) |>
  mutate(pct = n / sum(n)) |>
  filter(high_revenue == "high") |>
  ggplot(aes(season, pct, fill = season)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#4E6E8E","#00693D","#FFB81C","#8B3A3A")) +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "% High Revenue by Season",
       x = NULL, y = "% High Revenue") +
  theme_minimal(base_size = 9)

p1 + p2
```

## RO2: Model Fitting 

```{r}
#| echo: true
#| code-fold: true
# Train/test split — stratified on outcome
set.seed(617)
retail_ml <- retail |> select(-location, -date, -revenue, -store_id, -product_id)
split <- initial_split(retail_ml, prop = 0.80, strata = high_revenue)
train <- training(split)
test  <- testing(split)

# Recipe
rec <- recipe(high_revenue ~ ., data = train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_zv(all_predictors())

# Random Forest specification
rf_spec <- rand_forest(trees = 300, mtry = 4, min_n = 20) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

rf_wf  <- workflow() |> add_recipe(rec) |> add_model(rf_spec)
rf_fit <- fit(rf_wf, data = train)
```

## RO2: Model Evaluation 

::::: columns
::: {.column width="50%"}
```{r}
test_preds <- augment(rf_fit, new_data = test)

# ROC curve
roc_data <- test_preds |>
  roc_curve(truth = high_revenue, .pred_high)

ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#00693D", linewidth = 1.5) +
  geom_abline(linetype = "dashed", color = "gray60") +
  labs(title = "ROC Curve — Random Forest",
       x = "1 - Specificity", y = "Sensitivity") +
  theme_minimal(base_size = 10)
```
:::

::: {.column width="50%"}
```{r}
auc_val <- test_preds |>
  roc_auc(truth = high_revenue, .pred_high)

cm <- conf_mat(test_preds, truth = high_revenue, estimate = .pred_class)

metrics_tbl <- test_preds |>
  metric_set(roc_auc, accuracy, sens, spec)(
    truth = high_revenue,
    estimate = .pred_class,
    .pred_high
  )

metrics_tbl |>
  select(.metric, .estimate) |>
  mutate(.estimate = round(.estimate, 3)) |>
  rename(Metric = .metric, Value = .estimate) |>
  gt() |>
  tab_header(title = "Test Set Metrics") |>
  tab_options(table.font.size = px(12))
```
:::
:::::

## RO2: Variable Importance 

::::: columns
::: {.column width="55%"}
```{r}
#| fig-height: 4.0
rf_fit |>
  extract_fit_parsnip() |>
  vip(
    num_features = 10,
    aesthetics = list(fill = "#00693D", color = "white", alpha = 0.85)
  ) +
  labs(title = "Variable Importance — Random Forest",
       subtitle = "Impurity-based importance") +
  theme_minimal(base_size = 10)
```
:::

::: {.column width="45%"}
**Key findings from VIP:**

- **Units sold** is the dominant predictor — volume drives revenue class more than discount level alone
- **Marketing spend** is the second strongest predictor, suggesting campaigns with more investment produce better outcomes
- **Discount percentage** contributes, but is secondary to volume effects
- **Product category** matters — Electronics and Furniture are stronger revenue drivers than Groceries or Clothing
:::
:::::

## RO2: Summary Findings 

::::: columns
::: {.column width="60%"}
**What the model tells us:**

1.  Discount percentage alone is not the primary driver of whether a transaction generates high revenue. Unit volume and marketing spend matter more.

2.  Transactions in Electronics and Furniture are much more likely to generate above-median revenue, regardless of discount level.

3.  A tuned Random Forest achieves a ROC AUC of approximately 0.76, meaning the model correctly ranks a true high-revenue transaction above a low-revenue one about 78% of the time.

4.  Deployed as an interactive Shiny app, the model lets users explore how combinations of these features drive revenue predictions in real time.
:::

::: {.column width="40%"}
**Business recommendation:**

> Prioritize marketing spend and volume-driving promotions over deep discount rates. High discounts reduce per-transaction margin without reliably moving the revenue class needle, especially in lower-margin categories like Groceries and Clothing.
:::
:::::

------------------------------------------------------------------------

# RO3 — Luis Berumen

## RO3: Do Discounts Increase Sales Revenue?

**Analytics Objective:** Examine the relationship between discount percentage and Sales Revenue using regression models. Units Sold, Product Category, Marketing Spend, season, day of week, and holiday status are also included to determine whether Discount Percentage is an important predictor of Sales Revenue.

::::: columns
::: {.column width="55%"}
### ML Methods Used

- OLS Linear Regression (baseline)
- Ridge Regression
- Lasso Regression
- **Tuned Elastic Net (primary)**
- 10-fold cross-validation for model comparison

**Evaluation Metrics:** RMSE, MAE, and R-squared
:::

::: {.column width="45%"}
```{r}
#| label: ro3-discount-revenue
#| fig-height: 4.5

retail |>
  ggplot(aes(discount_pct, revenue)) +
  geom_jitter(
    alpha = 0.10,
    width = 0.25,
    color = "#23351F"
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "#D9A441",
    linewidth = 1.2
  ) +
  scale_x_continuous(
    labels = label_percent(scale = 1)
  ) +
  scale_y_continuous(
    labels = label_dollar()
  ) +
  labs(
    title = "Discount % vs. Sales Revenue",
    x = "Discount %",
    y = "Sales Revenue (USD)"
  ) +
  theme_minimal(base_size = 13)
```
:::
:::::

::: notes
RO3 asks whether discounts increase sales revenue. Discount percentage is the main variable of interest and sales revenue is the outcome. Units sold, product category, marketing spend, and other available variables are included because they may also be related to revenue. Regression models are used to examine these variables together and determine how useful discount percentage is for predicting sales revenue.
:::

## RO3: Data Wrangling & EDA 

::::: columns
::: {.column width="55%"}
```{r}
#| label: ro3-category-discount
#| fig-height: 5

retail |>
  group_by(category, discount_group) |>
  summarise(
    avg_revenue = mean(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(
    aes(
      category,
      avg_revenue,
      fill = discount_group
    )
  ) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c(
      "#4E6E8E",
      "#FFB81C",
      "#00693D"
    )
  ) +
  scale_y_continuous(
    labels = label_dollar()
  ) +
  labs(
    title = "Average Revenue by Category & Discount",
    x = NULL,
    y = "Average Revenue",
    fill = "Discount"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```
:::

::: {.column width="45%"}
```{r}
#| label: ro3-discount-summary

retail |>
  group_by(discount_group) |>
  summarise(
    Transactions = n(),
    `Avg Revenue` = mean(revenue, na.rm = TRUE),
    `Total Revenue` = sum(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    `Avg Revenue` = dollar(`Avg Revenue`),
    `Total Revenue` = dollar(`Total Revenue`)
  ) |>
  gt() |>
  tab_header(
    title = "Revenue by Discount Group"
  ) |>
  tab_options(
    table.font.size = px(16)
  )
```
:::
:::::

::: notes
The first part of RO3 looks at the relationship between discounts and revenue before fitting the models. The scatterplot shows the overall relationship between discount percentage and sales revenue. The category comparison and summary table show whether average revenue changes across discount levels and product categories.
:::

## RO3: Model Fitting 

```{r}
#| label: ro3-model-fitting
#| echo: true
#| code-fold: true
#| cache: true

set.seed(617)

ro3_data <-
  retail |>
  select(
    revenue,
    discount_pct,
    units,
    mktg_spend,
    category,
    day,
    holiday,
    season
  )

ro3_split <-
  initial_split(
    ro3_data,
    prop = 0.80,
    strata = revenue
  )

ro3_train <- training(ro3_split)
ro3_test <- testing(ro3_split)

ro3_rec <-
  recipe(
    revenue ~ .,
    data = ro3_train
  ) |>
  step_impute_median(
    all_numeric_predictors()
  ) |>
  step_novel(
    all_nominal_predictors()
  ) |>
  step_unknown(
    all_nominal_predictors()
  ) |>
  step_dummy(
    all_nominal_predictors()
  ) |>
  step_normalize(
    all_numeric_predictors()
  ) |>
  step_zv(
    all_predictors()
  )

set.seed(617)

ro3_folds <-
  vfold_cv(
    ro3_train,
    v = 10,
    strata = revenue
  )

ro3_ols_wf <-
  workflow() |>
  add_recipe(ro3_rec) |>
  add_model(
    linear_reg() |>
      set_engine("lm")
  )

ro3_ridge_wf <-
  workflow() |>
  add_recipe(ro3_rec) |>
  add_model(
    linear_reg(
      penalty = 0.1,
      mixture = 0
    ) |>
      set_engine("glmnet")
  )

ro3_lasso_wf <-
  workflow() |>
  add_recipe(ro3_rec) |>
  add_model(
    linear_reg(
      penalty = 0.1,
      mixture = 1
    ) |>
      set_engine("glmnet")
  )

ro3_en_spec <-
  linear_reg(
    penalty = tune(),
    mixture = tune()
  ) |>
  set_engine("glmnet")

ro3_en_wf <-
  workflow() |>
  add_recipe(ro3_rec) |>
  add_model(ro3_en_spec)

ro3_grid <-
  grid_regular(
    penalty(range = c(-4, 0)),
    mixture(),
    levels = c(8, 5)
  )

ro3_en_tuned <-
  tune_grid(
    ro3_en_wf,
    resamples = ro3_folds,
    grid = ro3_grid,
    metrics = metric_set(
      rmse,
      mae,
      rsq
    )
  )

ro3_best <-
  select_best(
    ro3_en_tuned,
    metric = "rmse"
  )

ro3_final_wf <-
  finalize_workflow(
    ro3_en_wf,
    ro3_best
  )

ro3_final_fit <-
  fit(
    ro3_final_wf,
    data = ro3_train
  )
```

::: notes
The data are split into 80% training data and 20% testing data. The training data are used to build and compare the models, while the testing data are saved for the final evaluation. OLS, Ridge, Lasso, and Elastic Net regression are compared. Elastic Net is tuned using 10-fold cross-validation to find the model settings that give the lowest prediction error.
:::

## RO3: Model Evaluation 

::::: columns
::: {.column width="55%"}
```{r}
#| label: ro3-cv-comparison
#| code-fold: true
#| cache: true

ro3_cv_metrics <-
  function(wf, label) {
    fit_resamples(
      wf,
      resamples = ro3_folds,
      metrics = metric_set(
        rmse,
        mae,
        rsq
      )
    ) |>
      collect_metrics() |>
      mutate(model = label)
  }

ro3_cv <-
  bind_rows(
    ro3_cv_metrics(
      ro3_ols_wf,
      "OLS"
    ),
    ro3_cv_metrics(
      ro3_ridge_wf,
      "Ridge"
    ),
    ro3_cv_metrics(
      ro3_lasso_wf,
      "Lasso"
    ),
    collect_metrics(ro3_en_tuned) |>
      semi_join(
        ro3_best,
        by = c(
          "penalty",
          "mixture"
        )
      ) |>
      mutate(
        model = "Tuned Elastic Net"
      )
  )

ro3_cv |>
  filter(.metric == "rmse") |>
  ggplot(
    aes(
      reorder(model, mean),
      mean
    )
  ) +
  geom_col(
    fill = "#23351F",
    alpha = 0.88
  ) +
  geom_errorbar(
    aes(
      ymin = mean - std_err,
      ymax = mean + std_err
    ),
    width = 0.2,
    color = "#D9A441"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = label_dollar()
  ) +
  labs(
    title = "10-Fold CV Model Comparison",
    x = NULL,
    y = "Mean RMSE"
  ) +
  theme_minimal(base_size = 12)
```
:::

::: {.column width="45%"}
```{r}
#| label: ro3-test-metrics
#| code-fold: true

ro3_preds <-
  augment(
    ro3_final_fit,
    new_data = ro3_test
  )

ro3_test_metrics <-
  ro3_preds |>
  metric_set(
    rmse,
    mae,
    rsq
  )(
    truth = revenue,
    estimate = .pred
  )

ro3_test_metrics |>
  mutate(
    Value = case_when(
      .metric %in% c("rmse", "mae") ~
        dollar(.estimate),
      TRUE ~
        as.character(
          round(.estimate, 3)
        )
    )
  ) |>
  transmute(
    Metric = toupper(.metric),
    Value
  ) |>
  gt() |>
  tab_header(
    title = "Final Test Metrics"
  ) |>
  tab_options(
    table.font.size = px(17)
  )
```
:::
:::::

::: notes
The models are compared to determine which one gives the best predictions for sales revenue. RMSE and MAE show the amount of prediction error, so lower values are better. R-squared shows how much of the variation in sales revenue is explained by the model, so a higher value is better. The final model is then evaluated using the testing data.
:::

## RO3: Feature Importance 

::::: columns
::: {.column width="55%"}
```{r}
#| label: ro3-vip
#| fig-height: 5.5
#| code-fold: true

ro3_importance <-
  ro3_final_fit |>
  extract_fit_parsnip() |>
  tidy() |>
  filter(
    term != "(Intercept)",
    estimate != 0
  ) |>
  mutate(
    importance = abs(estimate)
  ) |>
  slice_max(
    importance,
    n = 10
  )

ro3_importance |>
  ggplot(
    aes(
      importance,
      reorder(term, importance)
    )
  ) +
  geom_col(
    fill = "#23351F",
    alpha = 0.88
  ) +
  labs(
    title = "Feature Importance — Tuned Elastic Net",
    subtitle = "Absolute standardized coefficients",
    x = "Importance",
    y = NULL
  ) +
  theme_minimal(base_size = 12)
```
:::

::: {.column width="45%"}
### How to Interpret RO3

- Larger bars show which variables are more important for predicting Sales Revenue.
- The importance of **Discount Percentage** shows whether discounts remain useful after the other predictors are included.
- Units Sold, Marketing Spend, and Product Category may provide stronger predictive information.
- These results describe predictive relationships and **do not establish causation**.
:::
:::::

## RO3: Summary Findings 

::::: columns
::: {.column width="60%"}
### What This AO Answers

1.  RO3 examines whether Discount Percentage is related to Sales Revenue.
2.  Discount Percentage is evaluated with Units Sold, Marketing Spend, Product Category, season, day, and holiday status.
3.  OLS, Ridge, Lasso, and Elastic Net are compared using 10-fold cross-validation.
4.  Feature importance evaluates how useful Discount Percentage is relative to the other predictors.
:::

::: {.column width="40%"}
### Business Recommendation

> Retailers should evaluate Discount Percentage together with Units Sold, Marketing Spend, Product Category, and timing rather than using discounts alone when making revenue decisions.

**Final interpretation:** The final RMSE, MAE, R-squared, and feature-importance results determine how well the model predicts Sales Revenue and how important Discount Percentage is relative to the other variables.
:::
:::::

------------------------------------------------------------------------

# RO4 — Tim Arana

## RO4: Predicting Sales Revenue

::::: columns
::: {.column width="52%"}
### Analytics Objective

Predict **Sales Revenue (USD)** and evaluate how **Product Category** and other retail characteristics contribute to revenue predictions.

**Machine Learning Task**

- Supervised learning
- Regression
- Outcome: Sales Revenue (USD)
- 80% training / 20% testing
- 10-fold cross-validation
:::

::: {.column width="48%"}
### Models Compared

1.  Lasso Regression
2.  Tuned Decision Tree
3.  Random Forest
4.  XGBoost

### Key Predictors

- Units Sold
- Discount Percentage
- Marketing Spend
- Product Category
- Store Location
- Day of the Week
- Holiday Effect
:::
:::::

## RO4: Model Comparison 

::::: columns
::: {.column width="48%"}
### 10-Fold CV Results

**XGBoost: 1,963.20**

Decision Tree: 1,973.09

Lasso Regression: 1,978.16

Random Forest: 2,056.74

*Lower RMSE = better prediction*
:::

::: {.column width="52%"}
### Best-Performing Model

**XGBoost** achieved the lowest average cross-validated RMSE.

- Only **9.89 RMSE** better than Decision Tree
- Difference was **less than one standard error**
- The two leading models therefore performed similarly

### Takeaway

**XGBoost ranked first, but its predictive advantage over the simpler Decision Tree was modest.**
:::
:::::

## RO4: Model Interpretation 

::::: columns
::: {.column width="50%"}
### Variable Importance & SHAP

- **Units Sold** was the dominant predictor.
- **Product Category** provided additional predictive value.
- **Groceries** and **Furniture** were among the strongest category features.
- Discount Percentage and Marketing Spend had smaller contributions.
:::

::: {.column width="50%"}
### Partial Dependence

- Predicted revenue generally increased as **Units Sold** increased.
- The relationship was **positive but nonlinear**.
- XGBoost captured changes and plateaus in predicted revenue.

### Takeaway

**Sales volume drives revenue predictions, while Product Category provides additional information.**
:::
:::::

## RO4: Findings & Recommendation 

::::: columns
::: {.column width="55%"}
### Key Findings

- **XGBoost** achieved the lowest cross-validated RMSE.
- Its advantage over the Decision Tree was modest.
- **Units Sold** was the dominant predictor of Sales Revenue.
- **Product Category** provided additional predictive value, especially Groceries and Furniture.
:::

::: {.column width="45%"}
### Business Recommendation

- Use **XGBoost** when predictive performance is the priority.
- Consider the **Decision Tree** when interpretability is more important.
- Monitor sales volume closely.
- Use **category-specific merchandising and promotional strategies**.

### Live Model

[Retail Sales Revenue Predictor](https://tarana7.shinyapps.io/W10-ML-Deployment/)
:::
:::::

------------------------------------------------------------------------

# RO5 — Mickyas Shawel

## RO5: Marketing Spend × Discount Interaction 

::::: columns
::: {.column width="50%"}

**Research Objective**

Determine whether marketing spend strengthens the effect of discount percentage on retail sales revenue.

**Final Model**

`Revenue ~ Discount + Marketing Spend + Discount × Marketing Spend`

**Key Result**

- Marketing Spend was positively associated with Revenue.
- The interaction term was positive, indicating that discounts perform better when supported by higher marketing investment.
- Discounts alone produced smaller gains than coordinated promotional campaigns.

:::

::: {.column width="50%"}

```{r}
#| fig-height: 3.8

retail |>
  ggplot(
    aes(
      mktg_spend,
      revenue,
      color = discount_pct
    )
  ) +
  geom_point(alpha = .35) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black"
  ) +
  scale_y_continuous(
    labels = scales::label_dollar()
  ) +
  labs(
    title = "Marketing Spend vs Revenue",
    x = "Marketing Spend",
    y = "Revenue",
    color = "Discount %"
  ) +
  theme_minimal(base_size = 10)
```

:::
:::::


## RO5: Interaction Analysis 

::::: columns
::: {.column width="50%"}

```{r}
#| fig-height: 3.6

retail |>
  mutate(
    spend_group = ntile(mktg_spend, 3),
    spend_group = factor(
      spend_group,
      labels = c(
        "Low",
        "Medium",
        "High"
      )
    )
  ) |>
  group_by(spend_group) |>
  summarise(
    avg_discount = mean(discount_pct),
    avg_revenue = mean(revenue),
    .groups = "drop"
  ) |>
  ggplot(
    aes(
      avg_discount,
      avg_revenue,
      color = spend_group
    )
  ) +
  geom_point(size = 3) +
  geom_line() +
  scale_y_continuous(
    labels = scales::label_dollar()
  ) +
  labs(
    title = "Average Revenue by Spend Level",
    x = "Average Discount (%)",
    y = "Average Revenue"
  ) +
  theme_minimal(base_size = 10)
```

:::

::: {.column width="50%"}

### Business Interpretation

- Higher marketing investment consistently generated higher average revenue.
- Promotions were most effective when marketing support increased alongside discounts.
- Coordinating advertising budgets with promotional pricing can improve revenue performance more than using discounts independently.

:::
:::::

------------------------------------------------------------------------

# Conclusion

## Key Findings 

::::: columns
::: {.column width="50%"}
**Across all research objectives, our analysis found:**

- **RO1:** Seasonal and holiday timing have a measurable but secondary effect on revenue — Random Forest outperforms Elastic Net regression, indicating seasonality interacts non-linearly with category and discount strategy rather than driving revenue on its own.

- **RO2:** Discount percentage is not the primary driver of revenue class. Unit volume and marketing spend have stronger predictive power. A Random Forest model achieves strong discrimination between high and low revenue transactions (AUC ≈ 0.76).

- **RO3:** Regression models evaluate Discount Percentage alongside Units Sold, Marketing Spend, Product Category, and timing variables, showing that discounts should be considered together with other retail factors rather than in isolation.

- **RO4:** XGBoost achieved the lowest cross-validated RMSE for predicting Sales Revenue. Units Sold was the dominant predictor, while Product Category, especially Groceries and Furniture, provided additional predictive value.

- **RO5:** Marketing Spend appears to interact with Discount Percentage in shaping Sales Revenue, suggesting that discount and marketing decisions should be coordinated rather than managed independently.
:::

::: {.column width="50%"}
```{r}
#| fig-height: 3.8
retail |>
  group_by(category, season) |>
  summarise(avg_rev = mean(revenue), .groups = "drop") |>
  ggplot(aes(season, avg_rev, color = category, group = category)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("#00693D","#FFB81C","#4E6E8E","#8B3A3A")) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Avg Revenue by Season & Category",
       x = NULL, y = "Avg Revenue (USD)", color = "Category") +
  theme_minimal(base_size = 10)
```
:::
:::::

## Recommendations for Businesses 

::: incremental
1.  **Rethink the discount-volume tradeoff.** Higher discounts increase units sold slightly but do not reliably push transactions into the high-revenue category. Margin protection should be prioritized over aggressive discounting in lower-margin categories.

2.  **Invest marketing spend strategically.** Marketing spend is one of the strongest predictors of above-median revenue outcomes. Reallocating budget toward higher-ROI categories and seasons produces measurable revenue lift.

3.  **Category-specific promotional strategy.** Electronics and Furniture generate above-median revenue regardless of discount level. Promotional spend in these categories amplifies existing revenue advantages. Groceries and Clothing require volume-first strategies to reach the same outcomes.

4.  **Use predictive models for promotional planning.** The deployed Shiny app demonstrates that ML models can translate feature inputs into actionable revenue probability estimates, enabling data-driven campaign design rather than trial-and-error discounting.
:::

## Limitations and Future Research

::::: columns
::: {.column width="50%"}
**Current limitations:**

- **Synthetic data:** All findings are based on a Kaggle dataset designed for analytics practice, not real transaction data. Results may not generalize to actual retail environments.
- **No causal inference:** Our models are predictive, not causal. We cannot establish that discounts cause revenue changes, only that they are associated with them.
- **Missing variables:** Customer-level features (loyalty, lifetime value, demographics) are absent and would likely improve model performance.
- **Static model:** The deployed model does not retrain on new data. Real-world drift in discount response patterns would degrade performance over time.
:::

::: {.column width="50%"}
**Directions for future research:**

- Collect real transaction data with customer-level identifiers to enable customer lifetime value modeling
- Apply causal inference methods (e.g., difference-in-differences, instrumental variables) to isolate the true causal effect of discounts on revenue
- Build a full MLOps pipeline using `vetiver` and `pins` to automate model retraining as new data arrives
- Explore segmentation-based discount optimization: different optimal discount rates for different customer and product segments
:::
:::::

------------------------------------------------------------------------

# Appendix

## Appendix

::::: columns
::: {.column width="50%"}
**GitHub Repository:**

<https://github.com/mjshawell/IBM-6540-Group-Project>

**Presentation Slides (GitHub Pages):**

\[To be updated after Pages is enabled\]

**Deployed Model (Shiny App — RO2):**

<https://k9wfby-aldo-naranjo.shinyapps.io/retail-revenue-predictor/>
:::

::: {.column width="50%"}
**Dataset:**

Kaggle Synthetic Retail Sales Dataset\
30,000 transactions \| 2022–2024\
4 product categories \| 11 variables

**Group Members:**

- Jake Evans
- Mickyas Shawel
- Aldo Naranjo
- Luis Berumen
- Tim Arana

**Course:** IBM 6540 — Applied Machine Learning for Marketing\
**Institution:** Cal Poly Pomona \| Summer 2026
:::
:::::

## Thank You

::: {style="text-align: center; padding-top: 60px;"}
**Group 1 — IBM 6540**

Jake Evans \| Mickyas Shawel \| Aldo Naranjo \| Luis Berumen \| Tim Arana

*Cal Poly Pomona \| Summer 2026*

------------------------------------------------------------------------

**Questions?**
:::
