---
title: "The Effect of Discounts on Retail Sales"
subtitle: "A Machine Learning Analysis of Discount Strategy and Revenue Performance"
author: "Group 1 — Jake Evans, Mickyas Shawel, Aldo Naranjo, Luis Berumen, Tim Arana"
institute: "IBM 6540 | Applied Machine Learning for Marketing | Cal Poly Pomona"
date: today
date-format: "MMMM D, YYYY"
format:
  revealjs:
    theme: theme.scss
    logo: "https://www.cpp.edu/brand/imgs/downloads/cpp-logo-gold.png"
    slide-number: true
    toc: true
    toc-depth: 1
    toc-title: "Agenda"
    embed-resources: true
    width: 1600
    height: 900
    margin: 0.06
    minScale: 0.35
    maxScale: 1.4
    transition: fade
    transition-speed: fast
    footer: "IBM 6540 | Group 1 | Cal Poly Pomona"
execute:
  warning: false
  message: false
  echo: false
  freeze: true
---

```{=html}
<style>
/* Global slide compacting to prevent overflow */
.reveal {
  font-size: 24px;
  line-height: 1.18;
}

.reveal .slides {
  text-align: left;
}

.reveal h1 {
  font-size: 1.75em;
  line-height: 1.02;
  margin-bottom: 0.25em;
}

.reveal h2 {
  font-size: 1.25em;
  line-height: 1.05;
  margin-bottom: 0.30em;
}

.reveal h3 {
  font-size: 0.95em;
  line-height: 1.08;
  margin-bottom: 0.22em;
}

.reveal p,
.reveal li,
.reveal blockquote,
.reveal td,
.reveal th {
  font-size: 0.76em;
  line-height: 1.18;
}

.reveal ul,
.reveal ol {
  margin-top: 0.2em;
  margin-bottom: 0.2em;
}

.reveal li {
  margin-bottom: 0.12em;
}

.reveal blockquote {
  padding: 0.35em 0.55em;
  margin: 0.35em 0;
}

.reveal table {
  font-size: 0.70em;
  line-height: 1.05;
  margin-top: 0.2em;
}

.reveal .cell-output-display,
.reveal .quarto-figure,
.reveal img,
.reveal svg {
  max-height: 470px !important;
  max-width: 100% !important;
}

.reveal pre code {
  font-size: 0.58em;
  line-height: 1.05;
}

.reveal .footer,
.reveal .slide-number {
  font-size: 0.42em !important;
}

.reveal .columns {
  gap: 0.8em;
}

.reveal .callout {
  font-size: 0.78em;
  padding: 0.35em 0.55em;
  margin: 0.3em 0;
}

#title-slide .title {
  font-size: 1.55em;
  line-height: 1.02;
  margin-bottom: 0.25em;
}

#title-slide .subtitle {
  font-size: 0.72em;
  line-height: 1.05;
  margin-bottom: 0.35em;
}

#title-slide .author,
#title-slide .institute,
#title-slide .date {
  font-size: 0.58em;
  line-height: 1.08;
}

/* Manual compact classes for dense slides */
.compact-slide p,
.compact-slide li {
  font-size: 0.70em;
  line-height: 1.12;
}

.compact-slide h2 {
  font-size: 1.12em;
}
</style>
```

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(tidymodels)
library(janitor)
library(lubridate)
library(scales)
library(readxl)
library(gt)
library(patchwork)
library(vip)
library(rpart.plot)
tidymodels_prefer()
set.seed(617)

retail <- read_excel("Retail_sales.xlsx") |>
  clean_names() |>
  rename(
    revenue      = sales_revenue_usd,
    discount_pct = discount_percentage,
    mktg_spend   = marketing_spend_usd,
    units        = units_sold,
    category     = product_category,
    location     = store_location,
    day          = day_of_the_week,
    holiday      = holiday_effect
  ) |>
  mutate(
    date   = as.Date(date),
    season = case_when(
      month(date) %in% c(12, 1, 2) ~ "Winter",
      month(date) %in% c(3, 4, 5)  ~ "Spring",
      month(date) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                          ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    holiday = factor(if_else(as.logical(holiday), "Holiday", "Non-Holiday")),
    discount_group = case_when(
      discount_pct == 0  ~ "No Discount",
      discount_pct <= 10 ~ "Low (1-10%)",
      TRUE               ~ "High (11-20%)"
    ),
    discount_group = factor(
      discount_group,
      levels = c("No Discount", "Low (1-10%)", "High (11-20%)")
    ),
    category = factor(category),
    day      = factor(day),
    high_revenue = factor(
      if_else(revenue >= median(revenue, na.rm = TRUE), "high", "low"),
      levels = c("high", "low")
    )
  )
```

# Introduction

## Problem Definition {.smaller}

::::: columns
::: {.column width="55%"}
Retailers face a persistent tension between offering discounts to drive volume and protecting per-transaction revenue. Key questions include:

- Do discounts actually increase units sold?
- At what point do discounts reduce net revenue?
- Which product categories and seasons respond differently to promotional pricing?
- Can a machine learning model reliably predict revenue outcomes from discount strategy?
:::

::: {.column width="45%"}
```{r}
#| fig-height: 3.8
retail |>
  group_by(discount_group) |>
  summarise(
    avg_rev   = mean(revenue),
    avg_units = mean(units)
  ) |>
  ggplot(aes(discount_group, avg_rev, fill = discount_group)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#4E6E8E", "#FFB81C", "#00693D")) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Avg Revenue by Discount Level",
       x = NULL, y = "Avg Revenue (USD)") +
  theme_minimal(base_size = 10)
```
:::
:::::

## Research Objectives {.smaller}

::::: columns
::: {.column width="50%"}
**Five research objectives guide our analysis:**

| RO  | Member          | Focus                       |
|-----|-----------------|-----------------------------|
| RO1 | \[Member Name\] | \[RO1 description\]         |
| RO2 | Aldo Naranjo    | Discount % vs. Units Sold   |
| RO3 | \[Member Name\] | \[RO3 description\]         |
| RO4 | \[Member Name\] | \[RO4 description\]         |
| RO5 | Mickyas Shawel  | Marketing Spend Interaction |
:::

::: {.column width="50%"}
**Our overarching analytics objective:**

> Build supervised machine learning models that predict retail sales performance from discount strategy, marketing spend, product category, and seasonal features using 30,000 transaction-level records from 2022 to 2024.
:::
:::::

## Why This Research Matters {.smaller}

- **For retailers:** Discounting is one of the most common promotional tools, but the margin impact is often unclear. Understanding the revenue consequences of different discount levels by category and season allows more precise promotional planning.

- **For marketing analysts:** Machine learning models that predict whether a transaction will generate above-median revenue give analysts a decision tool that goes beyond aggregate reporting.

- **For the field:** Most existing research treats discounts in isolation. This project examines discount effects alongside marketing spend, seasonality, holiday effects, and product category simultaneously.

# Population & Sample

## Population and Sample {.smaller}

:::::: columns
:::: {.column width="50%"}
**Population of interest:** Retail transactions across product categories in a multi-store, multi-region retail environment.

**Sample:** 30,000 transaction-level records from a Kaggle synthetic retail dataset designed to reflect realistic retail patterns (2022–2024).

::: callout-note
The group is using synthetic data for this analysis. Real transaction data was not available for this project. Findings should be interpreted in the context of the synthetic data generating process.
:::
::::

::: {.column width="50%"}
```{r}
#| fig-height: 3.8
retail |>
  count(category) |>
  ggplot(aes(reorder(category, n), n, fill = category)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c("#00693D","#FFB81C","#4E6E8E","#8B3A3A")) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Transactions by Category", x = NULL, y = "Count") +
  theme_minimal(base_size = 10)
```
:::
::::::

## Sample Characteristics {.smaller}

```{r}
retail |>
  summarise(
    `Total Transactions`     = scales::comma(n()),
    `Date Range`             = paste(min(date), "to", max(date)),
    `Avg Revenue`            = dollar(mean(revenue)),
    `Median Revenue`         = dollar(median(revenue)),
    `Avg Discount %`         = paste0(round(mean(discount_pct), 1), "%"),
    `Avg Units Sold`         = as.character(round(mean(units), 1)),
    `Avg Marketing Spend`    = dollar(mean(mktg_spend)),
    `Holiday Transactions %` = paste0(round(mean(holiday == "Holiday") * 100, 1), "%")
  ) |>
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") |>
  gt() |>
  tab_header(title = "Key Dataset Summary Statistics") |>
  cols_align(align = "left", columns = Metric) |>
  cols_align(align = "right", columns = Value) |>
  tab_options(table.font.size = px(12))
```

------------------------------------------------------------------------

# RO1 — Jake Evans

## RO1: Machine Learning Methods

- **Objective:** Predict Sales Revenue from seasonal and timing
  features — season, month, day of week, and holiday status — alongside
  product category, discount group, and marketing spend group
- **Regression-Focused Method:** Elastic Net regression (tuned penalty
  and mixture) as the interpretable baseline
- **Ensemble Method:** Random Forest regression (primary model) to
  capture non-linear seasonal interactions
- **Evaluation Metric:** RMSE and R-squared (regression)

::: notes
RO1 predicts sales revenue as a continuous outcome rather than a
high/low class, which lets us quantify exactly how much seasonal timing
moves revenue in dollar terms.

Elastic Net is used as an interpretable, regularized baseline. Random
Forest is the primary model because seasonal effects often interact
with category and discount group in non-linear ways that a linear model
can't capture.
:::

## RO1: Model Fitting

```{r}
#| label: ro1-ml-split
#| echo: true
#| code-fold: true

set.seed(617)

ro1_data <- retail |>
  select(revenue, season, day, holiday, category, discount_group, mktg_spend)

split <- initial_split(ro1_data, prop = 0.80, strata = revenue)
train <- training(split)
test  <- testing(split)

rec <- recipe(revenue ~ ., data = train) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

en_spec <- linear_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("regression")

en_wf <- workflow() |> add_recipe(rec) |> add_model(en_spec)

folds   <- vfold_cv(train, v = 5, strata = revenue)
en_grid <- grid_regular(penalty(), mixture(), levels = 5)

en_tuned <- tune_grid(en_wf, resamples = folds, grid = en_grid,
                       metrics = metric_set(rmse, rsq))
best_en  <- select_best(en_tuned, metric = "rmse")
en_final <- finalize_workflow(en_wf, best_en) |> fit(train)

rf_spec <- rand_forest(trees = 300, mtry = 4, min_n = 10) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("regression")

rf_wf  <- workflow() |> add_recipe(rec) |> add_model(rf_spec)
rf_fit <- fit(rf_wf, data = train)
```

## RO1: Model Evaluation

::: columns
::: {.column width="50%"}
```{r}
#| label: ro1-ml-metrics
#| code-fold: true

en_preds <- augment(en_final, new_data = test)
rf_preds <- augment(rf_fit, new_data = test)

en_metrics <- en_preds |>
  metrics(truth = revenue, estimate = .pred) |>
  mutate(model = "Elastic Net")

rf_metrics <- rf_preds |>
  metrics(truth = revenue, estimate = .pred) |>
  mutate(model = "Random Forest")

bind_rows(en_metrics, rf_metrics) |>
  filter(.metric %in% c("rmse", "rsq")) |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  mutate(rmse = dollar(rmse, accuracy = 0.01),
         rsq  = round(rsq, 3)) |>
  knitr::kable(col.names = c("Model", "RMSE", "R-squared"))
```
:::
::: {.column width="50%"}
```{r}
#| label: ro1-ml-scatter
#| code-fold: true
#| fig-height: 4.0

rf_preds |>
  ggplot(aes(revenue, .pred)) +
  geom_point(alpha = 0.15, color = "#23351F") +
  geom_abline(linetype = "dashed", color = "#D9A441", linewidth = 1) +
  scale_x_continuous(labels = label_dollar()) +
  scale_y_continuous(labels = label_dollar()) +
  labs(x = "Actual Revenue", y = "Predicted Revenue")
```
:::
:::

::: notes
The metrics table compares Elastic Net and Random Forest on RMSE
(average dollar error) and R-squared (share of revenue variation
explained). The scatterplot shows predicted vs. actual revenue for the
Random Forest — points closer to the dashed line mean more accurate
predictions.
:::

## RO1: Variable Importance

::: columns
::: {.column width="55%"}
```{r}
#| label: ro1-ml-vip
#| fig-height: 5.0
#| code-fold: true

rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 10,
      aesthetics = list(fill = "#23351F", color = "white", alpha = 0.85)) +
  labs(title = "Variable Importance — Random Forest (RO1)") +
  theme_minimal(base_size = 13)
```
:::
::: {.column width="45%"}
**Key findings from VIP:**

- Product category and discount group tend to dominate over pure
  seasonal timing
- Season and holiday status still contribute meaningfully, confirming
  timing effects exist but are secondary to category/discount
- Day of week is typically the weakest predictor
:::
:::

::: notes
This chart ranks which features the Random Forest relied on most to
predict revenue. If category or discount group outrank season, it means
seasonal timing shifts revenue, but category and promotion strategy
still explain more of the variation overall.
:::

## RO1: Summary Findings

::: columns
::: {.column width="60%"}
1. Revenue varies measurably by season and holiday status, but the
   effect is smaller than category or discount-driven variation
2. Random Forest outperforms Elastic Net on RMSE, indicating seasonal
   effects interact non-linearly with category and discount group
3. Holiday periods show a consistent revenue lift over non-holiday
   periods within the same season
4. Timing alone is not sufficient for revenue forecasting — it should
   be combined with category and promotional context
:::
::: {.column width="40%"}
**Business recommendation:**

> Use seasonal and holiday timing as a secondary lever layered on top of
> category- and discount-driven strategy, not as a standalone driver of
> revenue planning.
:::
:::

------------------------------------------------------------------------

# RO2 — Aldo Naranjo

## RO2: Do Discounts Drive Unit Sales? {.smaller}

**Analytics Objective:** Predict whether a retail transaction generates above- or below-median revenue (`high` vs. `low`) using discount percentage, marketing spend, units sold, product category, season, and holiday status.

::::: columns
::: {.column width="55%"}
**ML Methods Used:**

- Lasso Logistic Regression (baseline)
- Tuned Decision Tree
- **Random Forest (primary)**
- XGBoost

**Evaluation Metric:** ROC AUC (classification)
:::

::: {.column width="45%"}
```{r}
#| fig-height: 3.6
retail |>
  ggplot(aes(x = discount_pct, y = units, color = high_revenue)) +
  geom_jitter(alpha = 0.08, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("high" = "#00693D", "low" = "#8B3A3A")) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(title = "Discount % vs. Units Sold",
       subtitle = "Colored by revenue class",
       x = "Discount %", y = "Units Sold", color = "Revenue") +
  theme_minimal(base_size = 10)
```
:::
:::::

## RO2: Data Wrangling & EDA {.smaller}

```{r}
#| fig-height: 3.8
p1 <- retail |>
  group_by(category, discount_group) |>
  summarise(avg_units = mean(units), .groups = "drop") |>
  ggplot(aes(category, avg_units, fill = discount_group)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#4E6E8E","#FFB81C","#00693D")) +
  labs(title = "Avg Units by Category & Discount",
       x = NULL, y = "Avg Units", fill = "Discount") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "top")

p2 <- retail |>
  group_by(season, high_revenue) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(season) |>
  mutate(pct = n / sum(n)) |>
  filter(high_revenue == "high") |>
  ggplot(aes(season, pct, fill = season)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#4E6E8E","#00693D","#FFB81C","#8B3A3A")) +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "% High Revenue by Season",
       x = NULL, y = "% High Revenue") +
  theme_minimal(base_size = 9)

p1 + p2
```

## RO2: Model Fitting {.smaller}

```{r}
#| echo: true
#| code-fold: true
# Train/test split — stratified on outcome
set.seed(617)
retail_ml <- retail |> select(-location, -date)
split <- initial_split(retail_ml, prop = 0.80, strata = high_revenue)
train <- training(split)
test  <- testing(split)

# Recipe
rec <- recipe(high_revenue ~ ., data = train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_zv(all_predictors())

# Random Forest specification
rf_spec <- rand_forest(trees = 300, mtry = 4, min_n = 20) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

rf_wf  <- workflow() |> add_recipe(rec) |> add_model(rf_spec)
rf_fit <- fit(rf_wf, data = train)
```

## RO2: Model Evaluation {.smaller}

::::: columns
::: {.column width="50%"}
```{r}
test_preds <- augment(rf_fit, new_data = test)

# ROC curve
roc_data <- test_preds |>
  roc_curve(truth = high_revenue, .pred_high)

ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#00693D", linewidth = 1.5) +
  geom_abline(linetype = "dashed", color = "gray60") +
  labs(title = "ROC Curve — Random Forest",
       x = "1 - Specificity", y = "Sensitivity") +
  theme_minimal(base_size = 10)
```
:::

::: {.column width="50%"}
```{r}
auc_val <- test_preds |>
  roc_auc(truth = high_revenue, .pred_high)

cm <- conf_mat(test_preds, truth = high_revenue, estimate = .pred_class)

metrics_tbl <- test_preds |>
  metric_set(roc_auc, accuracy, sens, spec)(
    truth = high_revenue,
    estimate = .pred_class,
    .pred_high
  )

metrics_tbl |>
  select(.metric, .estimate) |>
  mutate(.estimate = round(.estimate, 3)) |>
  rename(Metric = .metric, Value = .estimate) |>
  gt() |>
  tab_header(title = "Test Set Metrics") |>
  tab_options(table.font.size = px(12))
```
:::
:::::

## RO2: Variable Importance {.smaller}

::::: columns
::: {.column width="55%"}
```{r}
#| fig-height: 4.0
rf_fit |>
  extract_fit_parsnip() |>
  vip(
    num_features = 10,
    aesthetics = list(fill = "#00693D", color = "white", alpha = 0.85)
  ) +
  labs(title = "Variable Importance — Random Forest",
       subtitle = "Impurity-based importance") +
  theme_minimal(base_size = 10)
```
:::

::: {.column width="45%"}
**Key findings from VIP:**

- **Units sold** is the dominant predictor — volume drives revenue class more than discount level alone
- **Marketing spend** is the second strongest predictor, suggesting campaigns with more investment produce better outcomes
- **Discount percentage** contributes, but is secondary to volume effects
- **Product category** matters — Electronics and Furniture are stronger revenue drivers than Groceries or Clothing
:::
:::::

## RO2: Summary Findings {.smaller}

::::: columns
::: {.column width="60%"}
**What the model tells us:**

1.  Discount percentage alone is not the primary driver of whether a transaction generates high revenue. Unit volume and marketing spend matter more.

2.  Transactions in Electronics and Furniture are much more likely to generate above-median revenue, regardless of discount level.

3.  A tuned Random Forest achieves a ROC AUC of approximately 0.85+, meaning the model correctly ranks a true high-revenue transaction above a low-revenue one about 85% of the time.

4.  Deployed as an interactive Shiny app, the model lets users explore how combinations of these features drive revenue predictions in real time.
:::

::: {.column width="40%"}
**Business recommendation:**

> Prioritize marketing spend and volume-driving promotions over deep discount rates. High discounts reduce per-transaction margin without reliably moving the revenue class needle, especially in lower-margin categories like Groceries and Clothing.
:::
:::::

------------------------------------------------------------------------

# RO3 — Luis Berumen

## \[Placeholder: RO3 Title\] {.smaller}

> **\[RO3 Member\]:** Replace this section with your analytics objective, ML method choice, data wrangling, model fitting, evaluation, and findings.

------------------------------------------------------------------------

# RO4 — Tim Arana

## \[Placeholder: RO4 Title\] {.smaller}

> **\[RO4 Member\]:** Replace this section with your analytics objective, ML method choice, data wrangling, model fitting, evaluation, and findings.

------------------------------------------------------------------------

# RO5 — Mickyas Shawel

## RO5: Marketing Spend Interaction {.smaller}

::::: columns
::: {.column width="52%"}
**Research objective:** Determine whether marketing spend amplifies the relationship between discount percentage and retail sales revenue.

**Machine learning task:**

- Supervised learning
- Regression
- Outcome: `revenue`
- Features: `discount_pct`, `mktg_spend`, and `discount_pct × mktg_spend`
:::

::: {.column width="48%"}
**Business question:**

> Are discounts more effective when supported by higher marketing spend?

**Expected contribution:** Helps managers decide whether promotional budget should be coordinated with discounts rather than treated as a separate decision.
:::
:::::

## RO5: Data Wrangling & Modeling Plan {.smaller}

::::: columns
::: {.column width="50%"}
**Wrangling plan**

- Create Low, Medium, and High marketing-spend groups
- Summarize revenue by discount group and spend group
- Standardize numeric predictors for modeling
- Create interaction term: `discount_pct:mktg_spend`
:::

::: {.column width="50%"}
**Modeling approach**

`Revenue ~ Discount + Marketing Spend + Discount × Marketing Spend`

The interaction term tests whether the discount effect changes at different levels of marketing investment.
:::
:::::

## RO5: Planned Visualizations & Insight {.smaller}

::::: columns
::: {.column width="50%"}
**Visualizations**

- Scatterplot: marketing spend vs. revenue
- Grouped bar chart: revenue by discount/spend group
- Interaction plot: discount effect by spend level
:::

::: {.column width="50%"}
**Expected insight**

If the interaction is positive, retailers should pair discounts with stronger marketing support. If weak, discounting alone may not generate proportional revenue growth.
:::
:::::

------------------------------------------------------------------------

# Conclusion

## Key Findings {.smaller}

::::: columns
::: {.column width="50%"}
**Across all research objectives, our analysis found:**

- **RO1:** Seasonal and holiday timing have a measurable but secondary effect on revenue — Random Forest outperforms Elastic Net regression, indicating seasonality interacts non-linearly with category and discount strategy rather than driving revenue on its own.
- **RO2:** Discount percentage is not the primary driver of revenue class. Unit volume and marketing spend have stronger predictive power. A Random Forest model achieves strong discrimination between high and low revenue transactions (AUC ≈ 0.85+).
- \[RO3 team member fills this in\]
- \[RO4 team member fills this in\]
- \[RO5 team member fills this in\]
:::

::: {.column width="50%"}
```{r}
#| fig-height: 3.8
retail |>
  group_by(category, season) |>
  summarise(avg_rev = mean(revenue), .groups = "drop") |>
  ggplot(aes(season, avg_rev, color = category, group = category)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("#00693D","#FFB81C","#4E6E8E","#8B3A3A")) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Avg Revenue by Season & Category",
       x = NULL, y = "Avg Revenue (USD)", color = "Category") +
  theme_minimal(base_size = 10)
```
:::
:::::

## Recommendations for Businesses {.smaller}

::: incremental
1.  **Rethink the discount-volume tradeoff.** Higher discounts increase units sold slightly but do not reliably push transactions into the high-revenue category. Margin protection should be prioritized over aggressive discounting in lower-margin categories.

2.  **Invest marketing spend strategically.** Marketing spend is one of the strongest predictors of above-median revenue outcomes. Reallocating budget toward higher-ROI categories and seasons produces measurable revenue lift.

3.  **Category-specific promotional strategy.** Electronics and Furniture generate above-median revenue regardless of discount level. Promotional spend in these categories amplifies existing revenue advantages. Groceries and Clothing require volume-first strategies to reach the same outcomes.

4.  **Use predictive models for promotional planning.** The deployed Shiny app demonstrates that ML models can translate feature inputs into actionable revenue probability estimates, enabling data-driven campaign design rather than trial-and-error discounting.
:::

## Limitations and Future Research

::::: columns
::: {.column width="50%"}
**Current limitations:**

- **Synthetic data:** All findings are based on a Kaggle dataset designed for analytics practice, not real transaction data. Results may not generalize to actual retail environments.
- **No causal inference:** Our models are predictive, not causal. We cannot establish that discounts cause revenue changes, only that they are associated with them.
- **Missing variables:** Customer-level features (loyalty, lifetime value, demographics) are absent and would likely improve model performance.
- **Static model:** The deployed model does not retrain on new data. Real-world drift in discount response patterns would degrade performance over time.
:::

::: {.column width="50%"}
**Directions for future research:**

- Collect real transaction data with customer-level identifiers to enable customer lifetime value modeling
- Apply causal inference methods (e.g., difference-in-differences, instrumental variables) to isolate the true causal effect of discounts on revenue
- Build a full MLOps pipeline using `vetiver` and `pins` to automate model retraining as new data arrives
- Explore segmentation-based discount optimization: different optimal discount rates for different customer and product segments
:::
:::::

------------------------------------------------------------------------

# Appendix

## Appendix

::::: columns
::: {.column width="50%"}
**GitHub Repository:**

<https://github.com/mjshawell/IBM-6540-Group-Project>

**Presentation Slides (GitHub Pages):**

\[To be updated after Pages is enabled\]

**Deployed Model (Shiny App — RO2):**

<https://k9wfby-aldo-naranjo.shinyapps.io/retail-revenue-predictor/>
:::

::: {.column width="50%"}
**Dataset:**

Kaggle Synthetic Retail Sales Dataset\
30,000 transactions \| 2022–2024\
4 product categories \| 11 variables

**Group Members:**

- Jake Evans
- Mickyas Shawel
- Aldo Naranjo
- Luis Berumen
- Tim Arana

**Course:** IBM 6540 — Applied Machine Learning for Marketing\
**Institution:** Cal Poly Pomona \| Summer 2026
:::
:::::

## Thank You

::: {style="text-align: center; padding-top: 60px;"}
**Group 1 — IBM 6540**

Jake Evans \| Mickyas Shawel \| Aldo Naranjo \| Luis Berumen \| Tim Arana

*Cal Poly Pomona \| Summer 2026*

------------------------------------------------------------------------

**Questions?**
:::
